### 2023-05-01 paper/acl2023.tex
paper

% This must be in the first 5 lines to tell arXiv to use pdfLaTeX, which is strongly recommended.
\pdfoutput=1
% In particular, the hyperref package requires pdfLaTeX in order to break URLs across lines.

\documentclass[11pt]{article}

% Remove the "review" option to generate the final version.
\usepackage[review]{ACL2023}

% Standard package includes
\usepackage{times}
\usepackage{latexsym}

% For proper rendering and hyphenation of words containing Latin characters (including in bib files)
\usepackage[T1]{fontenc}
% For Vietnamese characters
% \usepackage[T5]{fontenc}
% See https://www.latex-project.org/help/documentation/encguide.pdf for other character sets

% This assumes your files are encoded as UTF8
\usepackage[utf8]{inputenc}

% This is not strictly necessary, and may be commented out.
% However, it will improve the layout of the manuscript,
% and will typically save some space.
\usepackage{microtype}

% This is also not strictly necessary, and may be commented out.
% However, it will improve the aesthetics of text in
% the typewriter font.
\usepackage{inconsolata}
\usepackage{graphicx}

% If the title and author information does not fit in the area allocated, uncomment the following
%
%\setlength\titlebox{<dim>}
%
% and set <dim> to something 5cm or larger.

\title{Instructions for ACL 2023 Proceedings}

\author{Leixin Zhang \\ 
    Computational Linguistics\\
    Seminar für Sprachwissenschaft,
    Universität Tübingen, Germany \\ 
    \texttt{leixin.zhang@student.uni-tuebingen.de}
    }%\\\And
  % Second Author \\
  % Affiliation / Address line 1 \\
  % Affiliation / Address line 2 \\
  % Affiliation / Address line 3 \\
  % \texttt{email@domain} \\}

\begin{document}
\maketitle
\begin{abstract}
This is abstract
\end{abstract}

\section{Introduction}

\section{Previous studies}

\section{Methodology}

\section{Evaluating Sentence Embeddings with Gaussian Mixture}

This section presents an unsupervised approach for evaluating the separability of sentence embeddings. We measure label separability pairwisely using a Gaussian Mixture model and calculate the F-score of the unsupervised clustered labels with the true labels.

The Gaussian Mixture model is a probabilistic model that assumes that each cluster follows a Gaussian (or normal) distribution and estimates the weight of the density function for each cluster \cite{reynolds2009gaussian,5298967}. We assume that the sentence vectors of two distinct classes should achieve high accuracy with Gaussian Mixture if they are displayed in a Gaussian distribution in space and are separable from each other.

However, there is a potential limitation to using this method. The separability and accuracy score may be underestimated if two clusters are not normally distributed, as illustrated in Figure \ref{fig:cirle}.

\begin{figure}[htp]
    \centering
    \includegraphics[scale=0.4]{Circle.jpg}
    \caption{Cases that Gaussian Mixture Underestimate the Separability of Two Clusters}
    \label{fig:cirle}
\end{figure}

Our evaluation results show that out of 78 pairs of derivation classes, 7 pairs achieved an accuracy score above 90\%, 24 pairs above 80\%, and 47 pairs above 60\%, as depicted in Figure \ref{fig:GM}.

In particular, the class `ban' demonstrates good separability with many other classes, achieving an accuracy score of 0.982 with `past', 0.980 with `formal sentence', 0.942 with `minimal change', and 0.937 with `future', among other pairs.

Additionally, our evaluation results reveal that tenses are generally well-separated, with an accuracy score of 0.90 for `past' and future' classes, and an accuracy score of 0.924 for `simple sentence' and `future' classes.

\begin{figure}[htp]
    \centering
    \includegraphics[scale=0.4]{GM.png}
    \caption{Accuracy of Measured Separability with Gaussian Mixture}
    \label{fig:GM}
\end{figure}

% \subsection{SBERT as a classifier}
% This section shows whether the sentence embeddings trained on our models can serve as a classifier and separate sentences in terms of labels.

% For each derivation sentence, we compute the distance from its seed. The resulting vector. It is assumed that a certain type of derivation from a seed shares some similarities. A good classifier is assumed to group the same derivation types. For example, (sentence vector - seed vector)sentences with future tense are supposed to group together.
% visualize the distance vecotors

% reduced dimension PCA

% Sentence embeddings trained with SBERT can serve as classifier for some labels. For instance, the embeddings can separate 'future','past' and 'simple sentence' as in Figure X
% However, it does not work for the rest of groups:

% \subsection{Degree of Transformation}

% Whether the degree of transformation can be characterized by SBERT sentence embeddings. 
% It shows the generalization group achieved most correlation.
% The more generalized sentences are further from seed sentence. While less generalized sentences (sentence meaning closer to seed sentences) has shorter distance to seed sentences

\bibliography{anthology}
\bibliographystyle{acl_natbib}

\end{document}

### 2023-05-01 src/sentence_topology/classification/analysis.py
Add option to specify scoring for cls grid search.
- Classifier analysis results now include score and therefore are
  backwards incompatible.

@@ -8,7 +8,7 @@ import matplotlib.gridspec as gridspec
 import matplotlib.pyplot as plt
 import numpy as np
 import pandas as pd
-from sklearn.metrics import classification_report, confusion_matrix
+from sklearn.metrics import classification_report, confusion_matrix, get_scorer
 from sklearn.model_selection import train_test_split
 from sklearn.preprocessing import LabelEncoder
 
@@ -54,7 +54,8 @@ class ClassifierAnalysisResults:
     classifier_type: type
     classifier_params: dict[str, Any]
     confusion_matrix: pd.DataFrame
-    accuracy: float
+    score: float
+    score_name: str
     macro_metrics: dict[str, float]
     report: pd.DataFrame
 
@@ -73,7 +74,7 @@ class ClassifierAnalysisResults:
             "\n".join(
                 [
                     f"classifier: {self.classifier_type.__name__}",
-                    f"accuracy: {self.accuracy:.5f}",
+                    f"{self.score_name}: {self.score:.5f}",
                     f"params: {self.classifier_params}",
                 ]
             ),
@@ -105,6 +106,7 @@ def analyze_classifier(
     classifier,
     *,
     test_split_size: float = 0.5,
+    scoring: str = "accuracy",
 ) -> ClassifierAnalysisResults:
     data = create_embedding_transformation_prediction_data(embeddings)
     class_names = cast(list[str], data.label_encoder.classes_)
@@ -141,11 +143,14 @@ def analyze_classifier(
         columns=class_names,
     )
 
+    score = get_scorer(scoring)(label_test, predictions)
+
     return ClassifierAnalysisResults(
         classifier_type=type(classifier),
         classifier_params=classifier.get_params(deep=False),
         report=report,
-        accuracy=cast(float, metrics["accuracy"]),
+        score=score,
+        score_name=scoring,
         macro_metrics=cast(dict[str, float], metrics["macro avg"]),
         confusion_matrix=conf_matrix,
     )

### 2023-05-01 src/sentence_topology/scripts/update_cls_gs.py

@@ -22,6 +22,8 @@ def update_grid_search_scores_and_params(
     scores: pd.DataFrame,
     params: pd.DataFrame,
     embeddings: list[tuple[str, list[CostraEmbedding]]],
+    *,
+    scoring: str,
 ) -> None:
     def should_update(name: str) -> bool:
         return name not in scores.index or name not in params.index
@@ -29,7 +31,9 @@ def update_grid_search_scores_and_params(
     for name, embedding in tqdm(embeddings):
         if should_update(name):
             evals = grid_search_classifiers_params(
-                embedding, DEFAULT_GRID_SEARCHED_CLASSIFIERS
+                embedding,
+                DEFAULT_GRID_SEARCHED_CLASSIFIERS,
+                scoring=scoring,
             )
             scores.loc[name] = [eval.best_score_ for eval in evals]
             params.loc[name] = [eval.best_params_ for eval in evals]
@@ -43,6 +47,7 @@ def update_analysis_results(
     analysis_results_dir: str,
     analysis_figs_dir: str,
     cls_type_by_name: dict[str, type],
+    scoring: str,
 ) -> None:
     for embed_name, embed_scores in tqdm(scores.iterrows(), total=scores.shape[0]):
         embed_name = str(embed_name)
@@ -58,7 +63,7 @@ def update_analysis_results(
             classifier = cls_type_by_name[best_classifier_name](**cls_params)
             embeddings = fetch_embedding(embed_name)
 
-            analysis = analyze_classifier(list(embeddings), classifier)
+            analysis = analyze_classifier(list(embeddings), classifier, scoring=scoring)
             analysis.classifier_params = params.loc[embed_name, best_classifier_name]
 
             analysis.save(analysis_save_path)
@@ -163,6 +168,12 @@ def parse_args() -> argparse.Namespace:
         ),
         required=True,
     )
+    parser.add_argument(
+        "--scoring",
+        type=str,
+        help="Scoring method to identify better classifiers.",
+        default="accuracy",
+    )
 
     args = parser.parse_args()
 
@@ -242,7 +253,9 @@ def main() -> None:
             if entry.name not in scores.index or entry.name not in params.index:
                 embeds_to_update.append((entry.name, fetch_embedding(entry.name)))
 
-        update_grid_search_scores_and_params(scores, params, embeds_to_update)
+        update_grid_search_scores_and_params(
+            scores, params, embeds_to_update, scoring=args.scoring
[…]

### 2023-05-03 src/sentence_topology/classification/analysis.py
Change interface for GridSearchClassifier.
- Instead of type store instance to allow for Pipeline estimators
- Use StandardScaler for kNN and SVC

@@ -13,8 +13,10 @@ from sklearn.model_selection import train_test_split
 from sklearn.preprocessing import LabelEncoder
 
 from sentence_topology.data_types import CostraEmbedding
-from sentence_topology.visualization.predictions import (draw_confusion_matrix,
-                                                         draw_distributions)
+from sentence_topology.visualization.predictions import (
+    draw_confusion_matrix,
+    draw_distributions,
+)
 
 
 @dataclass
@@ -143,7 +145,7 @@ def analyze_classifier(
         columns=class_names,
     )
 
-    score = get_scorer(scoring)(label_test, predictions)
+    score = get_scorer(scoring)(classifier, feat_test, label_test)
 
     return ClassifierAnalysisResults(
         classifier_type=type(classifier),

### 2023-05-03 src/sentence_topology/classification/grid_search.py

@@ -2,36 +2,39 @@ from dataclasses import dataclass
 from typing import Any
 
 import joblib
+from sklearn.base import BaseEstimator
 from sklearn.ensemble import RandomForestClassifier
 from sklearn.model_selection import GridSearchCV, StratifiedGroupKFold
 from sklearn.neighbors import KNeighborsClassifier
 from sklearn.neural_network._multilayer_perceptron import MLPClassifier
+from sklearn.pipeline import Pipeline
 from sklearn.svm import SVC
 from sklearn.tree import DecisionTreeClassifier
+from sklearn.preprocessing import StandardScaler
 from tqdm.auto import tqdm
 
-from sentence_topology.classification.analysis import \
-    create_embedding_transformation_prediction_data
+from sentence_topology.classification.analysis import (
+    create_embedding_transformation_prediction_data,
+)
 from sentence_topology.data_types import CostraEmbedding
-from sentence_topology.utils import load_all_embeddings
 
 
 @dataclass
 class GridSearchClassifier:
-    type_: type
+    estimator: BaseEstimator
     params: dict[str, list[Any]]
 
 
 @dataclass
 class EvaluatedGridSearchClassifier:
-    classifier_type: type
+    estimator: BaseEstimator
     best_params_: dict[str, Any]
     best_score_: float
 
 
 def grid_search_classifiers_params(
     embeddings: list[CostraEmbedding],
-    clsfiers: list[GridSearchClassifier],
+    clsfiers: dict[str, GridSearchClassifier],
     *,
     with_bar: bool = True,
     verbose: int = 0,
@@ -42,12 +45,12 @@ def grid_search_classifiers_params(
 
     evaluated = []
     with tqdm(
-        clsfiers, desc="Classifiers evaluated", disable=not with_bar
+        clsfiers.items(), desc="Classifiers evaluated", disable=not with_bar
     ) as progress_bar:
-        for classifier in progress_bar:
-            progress_bar.set_postfix(classifier=classifier.type_.__name__)
+        for cls_name, classifier in progress_bar:
+            progress_bar.set_postfix(classifier=cls_name)
             gs = GridSearchCV(
-                classifier.type_(),
+                classifier.estimator,
                 classifier.params,
                 scoring=scoring,
                 cv=StratifiedGroupKFold(),
@@ -65,7 +68,7 @@ def grid_search_classifiers_params(
 
             evaluated.append(
[…]

### 2023-05-03 src/sentence_topology/scripts/update_cls_gs.py

@@ -7,13 +7,17 @@ import pandas as pd
 from tqdm.auto import tqdm
 
 from sentence_topology.classification.analysis import (
-    ClassifierAnalysisResults, analyze_classifier)
+    ClassifierAnalysisResults,
+    analyze_classifier,
+)
 from sentence_topology.classification.grid_search import (
-    DEFAULT_GRID_SEARCHED_CLASSIFIERS, grid_search_classifiers_params)
+    DEFAULT_GRID_SEARCHED_CLASSIFIERS,
+    GridSearchClassifier,
+    grid_search_classifiers_params,
+)
 from sentence_topology.data_types import CostraEmbedding
 from sentence_topology.utils.io import load_embedding
-from sentence_topology.utils.transform import (CONTEXT_MODES,
-                                               contextualize_embeddings)
+from sentence_topology.utils.transform import CONTEXT_MODES, contextualize_embeddings
 
 logger = logging.getLogger(__name__)
 
@@ -25,57 +29,47 @@ def update_grid_search_scores_and_params(
     *,
     scoring: str,
 ) -> None:
-    def should_update(name: str) -> bool:
-        return name not in scores.index or name not in params.index
-
     for name, embedding in tqdm(embeddings):
-        if should_update(name):
-            evals = grid_search_classifiers_params(
-                embedding,
-                DEFAULT_GRID_SEARCHED_CLASSIFIERS,
-                scoring=scoring,
-            )
-            scores.loc[name] = [eval.best_score_ for eval in evals]
-            params.loc[name] = [eval.best_params_ for eval in evals]
+        evals = grid_search_classifiers_params(
+            embedding,
+            DEFAULT_GRID_SEARCHED_CLASSIFIERS,
+            scoring=scoring,
+        )
+        scores.loc[name] = [eval.best_score_ for eval in evals]
+        params.loc[name] = [eval.best_params_ for eval in evals]
 
 
 def update_analysis_results(
+    *,
     scores: pd.DataFrame,
     params: pd.DataFrame,
-    fetch_embedding: Callable[[str], list[CostraEmbedding]],
-    *,
+    embeddings: list[tuple[str, list[CostraEmbedding]]],
     analysis_results_dir: str,
     analysis_figs_dir: str,
-    cls_type_by_name: dict[str, type],
+    gs_cls_map: dict[str, GridSearchClassifier],
     scoring: str,
 ) -> None:
[…]

### 2023-05-03 src/sentence_topology/classification/analysis.py
Add option to equalize label distribution for transformation prediction.
- Implement EmbeddingsLoader to support all transformations in a
  convenient interface

@@ -13,10 +13,8 @@ from sklearn.model_selection import train_test_split
 from sklearn.preprocessing import LabelEncoder
 
 from sentence_topology.data_types import CostraEmbedding
-from sentence_topology.visualization.predictions import (
-    draw_confusion_matrix,
-    draw_distributions,
-)
+from sentence_topology.visualization.predictions import (draw_confusion_matrix,
+                                                         draw_distributions)
 
 
 @dataclass
@@ -109,6 +107,7 @@ def analyze_classifier(
     *,
     test_split_size: float = 0.5,
     scoring: str = "accuracy",
+    # TODO: do CV and aggregate results
 ) -> ClassifierAnalysisResults:
     data = create_embedding_transformation_prediction_data(embeddings)
     class_names = cast(list[str], data.label_encoder.classes_)

### 2023-05-03 src/sentence_topology/classification/grid_search.py

@@ -8,14 +8,13 @@ from sklearn.model_selection import GridSearchCV, StratifiedGroupKFold
 from sklearn.neighbors import KNeighborsClassifier
 from sklearn.neural_network._multilayer_perceptron import MLPClassifier
 from sklearn.pipeline import Pipeline
+from sklearn.preprocessing import StandardScaler
 from sklearn.svm import SVC
 from sklearn.tree import DecisionTreeClassifier
-from sklearn.preprocessing import StandardScaler
 from tqdm.auto import tqdm
 
-from sentence_topology.classification.analysis import (
-    create_embedding_transformation_prediction_data,
-)
+from sentence_topology.classification.analysis import \
+    create_embedding_transformation_prediction_data
 from sentence_topology.data_types import CostraEmbedding
 
 

### 2023-05-03 src/sentence_topology/scripts/update_cls_gs.py

@@ -1,23 +1,16 @@
 import argparse
 import logging
 import os
-from typing import Callable, Iterable
+from typing import Iterable
 
 import pandas as pd
 from tqdm.auto import tqdm
 
-from sentence_topology.classification.analysis import (
-    ClassifierAnalysisResults,
-    analyze_classifier,
-)
+from sentence_topology.classification.analysis import analyze_classifier
 from sentence_topology.classification.grid_search import (
-    DEFAULT_GRID_SEARCHED_CLASSIFIERS,
-    GridSearchClassifier,
-    grid_search_classifiers_params,
-)
-from sentence_topology.data_types import CostraEmbedding
-from sentence_topology.utils.io import load_embedding
-from sentence_topology.utils.transform import CONTEXT_MODES, contextualize_embeddings
+    DEFAULT_GRID_SEARCHED_CLASSIFIERS, GridSearchClassifier,
+    grid_search_classifiers_params)
+from sentence_topology.utils.transform import CONTEXT_MODES, EmbeddingsLoader
 
 logger = logging.getLogger(__name__)
 
@@ -25,13 +18,15 @@ logger = logging.getLogger(__name__)
 def update_grid_search_scores_and_params(
     scores: pd.DataFrame,
     params: pd.DataFrame,
-    embeddings: list[tuple[str, list[CostraEmbedding]]],
+    embed_loader: EmbeddingsLoader,
     *,
     scoring: str,
 ) -> None:
-    for name, embedding in tqdm(embeddings):
+    for name in embed_loader.list_all(
+        tqdm_enable=True, tqdm_desc="Grid searching embeddings"
+    ):
         evals = grid_search_classifiers_params(
-            embedding,
+            embed_loader.load(name),
             DEFAULT_GRID_SEARCHED_CLASSIFIERS,
             scoring=scoring,
         )
@@ -43,13 +38,15 @@ def update_analysis_results(
     *,
     scores: pd.DataFrame,
     params: pd.DataFrame,
-    embeddings: list[tuple[str, list[CostraEmbedding]]],
+    embed_loader: EmbeddingsLoader,
     analysis_results_dir: str,
     analysis_figs_dir: str,
     gs_cls_map: dict[str, GridSearchClassifier],
     scoring: str,
 ) -> None:
-    for embed_name, embedding in tqdm(embeddings):
+    for embed_name in embed_loader.list_all(
[…]

### 2023-05-03 src/sentence_topology/utils/transform.py

@@ -1,6 +1,21 @@
+import logging
+import os
+from glob import glob
+from random import Random
+from typing import Callable, Iterable, Optional
+
 import numpy as np
+from tqdm.auto import tqdm
 
 from sentence_topology.data_types import CostraEmbedding
+from sentence_topology.utils.io import load_embedding
+
+logger = logging.getLogger(__name__)
+
+
+class EmbeddingContextualizationError(Exception):
+    pass
+
 
 CONTEXT_MODES = {
     "diff": lambda seed, sent: sent - seed,
@@ -12,20 +27,19 @@ def contextualize_embeddings(
     embeddings: list[CostraEmbedding],
     *,
     mode: str = "diff",
-) -> tuple[list[CostraEmbedding], int]:
+) -> list[CostraEmbedding]:
     id_to_embed = {}
 
     for embed in embeddings:
         if embed.trans == "seed":
             id_to_embed[embed.seed_id] = embed.embedding
 
-    skipped_count = 0
     transform = CONTEXT_MODES[mode]
     transformed = []
     for embed in embeddings:
         if embed.seed_id not in id_to_embed:
-            skipped_count += 1
-            continue
+            raise EmbeddingContextualizationError()
+
         new_embedding = transform(id_to_embed[embed.seed_id], embed.embedding)
         transformed.append(
             CostraEmbedding(
@@ -36,4 +50,80 @@ def contextualize_embeddings(
             )
         )
 
-    return transformed, skipped_count
+    return transformed
+
+
+def equalize_transformations(
+    embeddings: list[CostraEmbedding],
+    *,
+    include_seed: bool = False,
+    random_state: Optional[int] = None,
+    shuffle_after: bool = True,
+) -> list[CostraEmbedding]:
+    by_trans = {}
+    for embed in embeddings:
+        if embed.trans not in by_trans:
+            by_trans[embed.trans] = []
+
+        by_trans[embed.trans].append(embed)
+
+    if not include_seed and "seed" in by_trans:
+        del by_trans["seed"]
+
+    rand_gen = Random(random_state)
+    samples_num = min(len(embeds) for embeds in by_trans.values())
+    new_embeddings = []
+    for embeds in by_trans.values():
+        for sampled_embed in rand_gen.sample(embeds, k=samples_num):
+            new_embeddings.append(sampled_embed)
+
+    if shuffle_after:
+        rand_gen.shuffle(new_embeddings)
+
+    return new_embeddings
[…]

### 2023-05-04 src/sentence_topology/classification/analysis.py
Add option to analyse classifier with CV.
- `update_cls_gs` to support this option
- richer ClassifierAnalysisResults
- ClassifierAnalysisResults can now aggregate multiple results into one
- distribution visualizations can now work with floats

@@ -2,6 +2,7 @@ from __future__ import annotations
 
 import pickle
 from dataclasses import dataclass
+from functools import reduce
 from typing import Any, cast
 
 import matplotlib.gridspec as gridspec
@@ -9,7 +10,7 @@ import matplotlib.pyplot as plt
 import numpy as np
 import pandas as pd
 from sklearn.metrics import classification_report, confusion_matrix, get_scorer
-from sklearn.model_selection import train_test_split
+from sklearn.model_selection import StratifiedGroupKFold, train_test_split
 from sklearn.preprocessing import LabelEncoder
 
 from sentence_topology.data_types import CostraEmbedding
@@ -56,7 +57,9 @@ class ClassifierAnalysisResults:
     confusion_matrix: pd.DataFrame
     score: float
     score_name: str
-    macro_metrics: dict[str, float]
+    macro_avg: dict[str, float]
+    weighted_avg: dict[str, float]
+    accuracy: float
     report: pd.DataFrame
 
     def save(self, path: str) -> None:
@@ -67,6 +70,121 @@ class ClassifierAnalysisResults:
     def load(cls, path: str) -> ClassifierAnalysisResults:
         with open(path, mode="rb") as save_file:
             return pickle.load(save_file)
+
+    @classmethod
+    def from_trained_classifier(
+        cls,
+        classifier,
+        test_inputs,
+        test_labels,
+        *,
+        scoring: str,
+        class_names: list[str],
+    ) -> ClassifierAnalysisResults:
+        predictions = classifier.predict(test_inputs)
+        metrics = cast(
+            dict[str, Any],
+            classification_report(
+                test_labels,
+                predictions,
+                target_names=class_names,
+                output_dict=True,
+            ),
+        )
+
+        rows = {}
+        for label_name in class_names:
+            rows[label_name] = metrics.pop(label_name)
+
+        report = pd.DataFrame(rows)
+
+        conf_matrix = confusion_matrix(test_labels, predictions)
+        conf_matrix = pd.DataFrame(
+            conf_matrix,
+            index=class_names,
+            columns=class_names,
+        )
+
+        score = get_scorer(scoring)(classifier, test_inputs, test_labels)
+
+        return cls(
+            classifier_type=type(classifier),
+            classifier_params=classifier.get_params(deep=False),
+            report=report,
+            score=score,
+            score_name=scoring,
[…]

### 2023-05-04 src/sentence_topology/scripts/update_cls_gs.py

@@ -25,6 +25,8 @@ def update_grid_search_scores_and_params(
     for name in embed_loader.list_all(
         tqdm_enable=True, tqdm_desc="Grid searching embeddings"
     ):
+        if name in scores.index and name in params.index:
+            continue
         evals = grid_search_classifiers_params(
             embed_loader.load(name),
             DEFAULT_GRID_SEARCHED_CLASSIFIERS,
@@ -43,6 +45,7 @@ def update_analysis_results(
     analysis_figs_dir: str,
     gs_cls_map: dict[str, GridSearchClassifier],
     scoring: str,
+    cross_validation: bool,
 ) -> None:
     for embed_name in embed_loader.list_all(
         tqdm_enable=True, tqdm_desc="Analysing embeddings"
@@ -57,7 +60,10 @@ def update_analysis_results(
         classifier.set_params(**cls_params)
 
         analysis = analyze_classifier(
-            embed_loader.load(embed_name), classifier, scoring=scoring
+            embed_loader.load(embed_name),
+            classifier,
+            scoring=scoring,
+            cross_validation=cross_validation,
         )
         analysis.classifier_params = cls_params
 
@@ -144,22 +150,29 @@ def parse_args() -> argparse.Namespace:
         ),
     )
     parser.add_argument(
+        "--cv_analysis",
+        type=bool,
+        action=argparse.BooleanOptionalAction,
+        help=(
+            "Run classifier analysis using cross-validation and aggregate the results."
+        ),
+        default=False,
+    )
+    parser.add_argument(
         "--gs_scores",
         type=str,
         help=(
             "Path to pickled DataFrame containing embedding-classifier table of highest"
-            " scores."
+            " scores. Defaults to `results_dir`/scores.pkl."
         ),
-        required=True,
     )
     parser.add_argument(
         "--gs_params",
         type=str,
         help=(
             "Path to pickled DataFrame containing embedding-classifier table of best"
-            " params."
+            " params. Defaults to `results_dir`/params.pkl"
         ),
-        required=True,
     )
     parser.add_argument(
         "--scoring",
@@ -198,6 +211,12 @@ def parse_args() -> argparse.Namespace:
     )
     os.makedirs(args.analysis_figs_dir, exist_ok=True)
 
+    if args.gs_scores is None:
+        args.gs_scores = os.path.join(args.results_dir, "scores.pkl")
+
[…]

### 2023-05-04 src/sentence_topology/utils/transform.py

@@ -90,12 +90,10 @@ class EmbeddingsLoader:
         *,
         context_mode: Optional[str],
         equalize_trans: bool,
-        filter_cb: Optional[Callable[[str], bool]] = None,
     ) -> None:
         self._embeddings_dir = embeddings_dir
         self._context_mode = context_mode
         self._equalize_trans = equalize_trans
-        self._filter = (lambda _: True) if filter_cb is None else filter_cb
 
     def list_all(
         self, *, tqdm_enable: bool = False, tqdm_desc: str = "Processing embeddings"
@@ -109,7 +107,7 @@ class EmbeddingsLoader:
         ) as progress_bar:
             for entry in progress_bar:
                 progress_bar.set_postfix(embedding=entry.name)
-                if entry.name.endswith(".tsv") and self._filter(entry.name):
+                if entry.name.endswith(".tsv"):
                     try:
                         self.load(entry.name)
                         yield entry.name

### 2023-05-04 src/sentence_topology/visualization/predictions.py

@@ -25,23 +25,7 @@ def draw_confusion_matrix(
 def draw_distributions(conf_matrix: pd.DataFrame, axes: plt.Axes) -> None:
     true = conf_matrix.sum(axis=1)
     pred = conf_matrix.sum(axis=0)
-    types = []
-    classes = []
-
-    for type_, counts in zip(["True", "Predicted"], [true, pred]):
-        for cls, count in counts.items():
-            for _ in range(count):
-                types.append(type_)
-                classes.append(cls)
-
-    dists = pd.DataFrame({"Distribution": types, "Classes": classes})
-    sns.histplot(
-        dists,
-        x="Classes",
-        hue="Distribution",
-        element="step",
-        stat="percent",
-        ax=axes,
-    )
-    axes.set_xticks(axes.get_xticks())
-    axes.set_xticklabels(axes.get_xticklabels(), rotation=90)
+    hists = pd.DataFrame({"Predicted": preds, "True": true, "Label": preds.index})
+    hists.set_index("Label", inplace=True)
+    hists.plot.bar(ax=axes)
+    axes.set_xticks(rotation=90)

### 2023-05-04 notebooks/tmp.sync.py
Debug notebook.

@@ -1,0 +1,115 @@
+from typing import Any, cast
+
+import pandas as pd
+from sklearn.metrics import classification_report, confusion_matrix, get_scorer
+from sklearn.model_selection import train_test_split
+from sklearn.tree import DecisionTreeClassifier
+
+from sentence_topology.classification.analysis import (
+    ClassifierAnalysisResults, create_embedding_transformation_prediction_data)
+from sentence_topology.utils.io import load_embedding
+from sentence_topology.utils.transform import equalize_transformations
+from sentence_topology.visualization.predictions import draw_distributions
+
+# %%
+
+
+embedding = list(
+    load_embedding("../embeddings/paraphrase-multilingual-MiniLM-L12-v2.tsv")
+)
+# %%
+embedding = equalize_transformations(embedding)
+# %%
+data = create_embedding_transformation_prediction_data(embedding)
+
+# %%
+scoring = "f1_macro"
+
+# %%
+classifier = DecisionTreeClassifier()
+
+feat_train, feat_test, label_train, label_test = train_test_split(
+    data.features,
+    data.labels,
+    test_size=0.5,
+    stratify=data.labels,
+)
+classifier.fit(feat_train, label_train)
+
+class_names = cast(list[str], data.label_encoder.classes_)
+
+predictions = classifier.predict(feat_test)
+metrics = cast(
+    dict[str, Any],
+    classification_report(
+        label_test,
+        predictions,
+        target_names=class_names,
+        output_dict=True,
+    ),
+)
+# %%
+print(metrics)
+# %%
+
+rows = {}
+for label_name in class_names:
+    rows[label_name] = metrics.pop(label_name)
+
+report = pd.DataFrame(rows)
+# %%
+report
+# %%
+metrics
+# %%
+
+conf_matrix = confusion_matrix(label_test, predictions)
+conf_matrix = pd.DataFrame(
+    conf_matrix,
+    index=class_names,
+    columns=class_names,
+)
+
+score = get_scorer(scoring)(classifier, feat_test, label_test)
+
+analysis = ClassifierAnalysisResults(
+    classifier_type=type(classifier),
+    classifier_params=classifier.get_params(deep=False),
+    report=report,
+    score=score,
+    score_name=scoring,
+    confusion_matrix=conf_matrix,
[…]

### 2023-05-04 src/sentence_topology/classification/analysis.py
Improve visualization for classifier results

@@ -1,6 +1,7 @@
 from __future__ import annotations
 
 import pickle
+import seaborn as sns
 from dataclasses import dataclass
 from functools import reduce
 from typing import Any, cast
@@ -14,8 +15,11 @@ from sklearn.model_selection import StratifiedGroupKFold, train_test_split
 from sklearn.preprocessing import LabelEncoder
 
 from sentence_topology.data_types import CostraEmbedding
-from sentence_topology.visualization.predictions import (draw_confusion_matrix,
-                                                         draw_distributions)
+from sentence_topology.visualization.predictions import (
+    draw_classification_report,
+    draw_confusion_matrix,
+    draw_distributions_from_conf_matrix,
+)
 
 
 @dataclass
@@ -187,34 +191,42 @@ class ClassifierAnalysisResults:
         return agg_dict
 
     def visualize(self) -> plt.Figure:
-        fig = plt.figure(figsize=(12, 14))
+        sns.set_theme()
+        fig = plt.figure(figsize=(12, 20))
         fig.suptitle(
             "\n".join(
                 [
                     f"classifier: {self.classifier_type.__name__}",
-                    f"{self.score_name}: {self.score:.5f}",
                     f"params: {self.classifier_params}",
+                    f"{self.score_name}: {self.score:.5f}",
+                    f"accuracy: {self.accuracy:.5f}",
+                    f"macro avg.: {self.macro_avg}",
+                    f"weighted avg.: {self.weighted_avg}",
                 ]
             ),
             x=0.44,
             y=0.95,
         )
         grid = gridspec.GridSpec(
-            2,
+            3,
             2,
             figure=fig,
+            top=0.87,
             width_ratios=[7.5, 1],
-            height_ratios=[8, 4],
+            height_ratios=[10, 4, 4],
             hspace=0.6,
         )
 
-        conf_matrix_axis = fig.add_subplot(grid[:-1, :])
-        dist_matrix_axis = fig.add_subplot(grid[-1, :-1])
+        conf_matrix_axis = fig.add_subplot(grid[0, :])
+        dist_axis = fig.add_subplot(grid[1, :-1])
+        report_axis = fig.add_subplot(grid[-1, :-1])
 
         conf_matrix_axis.set_title("Normalized confusion matrix")
         draw_confusion_matrix(self.confusion_matrix, conf_matrix_axis)
-        dist_matrix_axis.set_title("Label distribution")
[…]

### 2023-05-04 src/sentence_topology/visualization/predictions.py

@@ -22,10 +22,22 @@ def draw_confusion_matrix(
     axes.set_xlabel("Predicted")
 
 
-def draw_distributions(conf_matrix: pd.DataFrame, axes: plt.Axes) -> None:
-    true = conf_matrix.sum(axis=1)
-    pred = conf_matrix.sum(axis=0)
-    hists = pd.DataFrame({"Predicted": preds, "True": true, "Label": preds.index})
-    hists.set_index("Label", inplace=True)
-    hists.plot.bar(ax=axes)
-    axes.set_xticks(rotation=90)
+def draw_distributions_from_conf_matrix(
+    conf_matrix: pd.DataFrame, axes: plt.Axes
+) -> None:
+    trues = conf_matrix.sum(axis=1)
+    preds = conf_matrix.sum(axis=0)
+    # hists = pd.DataFrame({"Predicted": preds, "True": trues})
+    # sns.lineplot(hists, ax=axes)
+    xs = range(len(trues.index))
+    axes.bar(xs, trues, label="True", alpha=0.4)
+    axes.bar(xs, preds, label="Predicted", alpha=0.4)
+    axes.legend()
+    axes.set_xticks(xs, labels=trues.index, rotation=90)
+
+
+def draw_classification_report(report: pd.DataFrame, axes: plt.Axes) -> None:
+    report = report.drop("support").stack().reset_index()
+    report.columns = ["Metric", "Label", "Value"]
+    sns.barplot(report, x="Label", y="Value", hue="Metric")
+    axes.set_xticks(axes.get_xticks(), labels=axes.get_xticklabels(), rotation=90)

### 2023-05-04 src/sentence_topology/classification/analysis.py
Figure spacing problem for classifier analysis results.

@@ -214,7 +214,7 @@ class ClassifierAnalysisResults:
             top=0.87,
             width_ratios=[7.5, 1],
             height_ratios=[10, 4, 4],
-            hspace=0.6,
+            hspace=0.65,
         )
 
         conf_matrix_axis = fig.add_subplot(grid[0, :])

### 2023-05-05 paper/paper.tex
Add Makefile for paper, reasonable file names, figure folder.

--- 2023-05-01 paper/acl2023.tex
@@
 
-This section presents an unsupervised approach for evaluating the separability of sentence embeddings.
-We measure label separability pairwisely using a Gaussian Mixture model and calculate the F-score of the unsupervised clustered labels with the true labels.
+This section presents an unsupervised approach for evaluating the separability
+of sentence embeddings. We measure label separability pairwisely using a
+Gaussian Mixture model and calculate the F-score of the unsupervised clustered
+labels with the true labels.
 
-The Gaussian Mixture model is a probabilistic model that assumes that each cluster follows a Gaussian (or normal) distribution and estimates the weight of the density function for each cluster \cite{reynolds2009gaussian,5298967}.
-We assume that the sentence vectors of two distinct classes should achieve high accuracy with Gaussian Mixture if they are displayed in a Gaussian distribution in space and are separable from each other.
+The Gaussian Mixture model is a probabilistic model that assumes that each
+cluster follows a Gaussian (or normal) distribution and estimates the weight of
+the density function for each cluster~\cite{reynolds2009gaussian,5298967}. We
+assume that the sentence vectors of two distinct classes should achieve high
+accuracy with Gaussian Mixture if they are displayed in a Gaussian distribution
+in space and are separable from each other.
 
-However, there is a potential limitation to using this method.
-The separability and accuracy score may be underestimated if two clusters are not normally distributed, as illustrated in Figure \ref{fig:cirle}.
+However, there is a potential limitation to using this method. The separability
+and accuracy score may be underestimated if two clusters are not normally
+distributed, as illustrated in Figure~\ref{fig:cirle}.
 
@@
     \centering
-    \includegraphics[scale=0.4]{Circle.jpg}
-    \caption{Cases that Gaussian Mixture Underestimate the Separability of Two Clusters}
-    \label{fig:cirle}
+    \includegraphics[scale=0.4]{figs/Circle.jpg}
+    \caption{Cases that Gaussian Mixture Underestimate the Separability of Two
+    Clusters}\label{fig:cirle}
 \end{figure}
 
-Our evaluation results show that out of 78 pairs of derivation classes, 7 pairs achieved an accuracy score above 90\%, 24 pairs above 80\%, and 47 pairs above 60\%, as depicted in Figure \ref{fig:GM}.
+Our evaluation results show that out of 78 pairs of derivation classes, 7 pairs
+achieved an accuracy score above 90\%, 24 pairs above 80\%, and 47 pairs above
+60\%, as depicted in Figure~\ref{fig:GM}.
 
-In particular, the class `ban' demonstrates good separability with many other classes, achieving an accuracy score of 0.982 with `past', 0.980 with `formal sentence', 0.942 with `minimal change', and 0.937 with `future', among other pairs.
+In particular, the class `ban' demonstrates good separability with many other
+classes, achieving an accuracy score of 0.982 with `past', 0.980 with `formal
+sentence', 0.942 with `minimal change', and 0.937 with `future', among other
+pairs.
 
-Additionally, our evaluation results reveal that tenses are generally well-separated, with an accuracy score of 0.90 for `past' and future' classes, and an accuracy score of 0.924 for `simple sentence' and `future' classes.
+Additionally, our evaluation results reveal that tenses are generally
+well-separated, with an accuracy score of 0.90 for `past' and future' classes,
+and an accuracy score of 0.924 for `simple sentence' and `future' classes.
 
@@
     \centering
-    \includegraphics[scale=0.4]{GM.png}
-    \caption{Accuracy of Measured Separability with Gaussian Mixture}
-    \label{fig:GM}
+    \includegraphics[scale=0.4]{figs/GM.png}
+    \caption{Accuracy of Measured Separability with Gaussian
+    Mixture}\label{fig:GM}
 \end{figure}
@@
 % \subsection{SBERT as a classifier}
-% This section shows whether the sentence embeddings trained on our models can serve as a classifier and separate sentences in terms of labels.
+% This section shows whether the sentence embeddings trained on our models can
+% serve as a classifier and separate sentences in terms of labels.
 
-% For each derivation sentence, we compute the distance from its seed.
-The resulting vector.
-It is assumed that a certain type of derivation from a seed shares some similarities.
-A good classifier is assumed to group the same derivation types.
-For example, (sentence vector - seed vector)sentences with future tense are supposed to group together.
+% For each derivation sentence, we compute the distance from its seed. The
+% resulting vector. It is assumed that a certain type of derivation from a seed
+% shares some similarities. A good classifier is assumed to group the same
+% derivation types. For example, (sentence vector - seed vector)sentences with
+% future tense are supposed to group together.
 % visualize the distance vecotors
@@
 
-% Sentence embeddings trained with SBERT can serve as classifier for some labels. For instance, the embeddings can separate 'future','past' and 'simple sentence' as in Figure X
+% Sentence embeddings trained with SBERT can serve as classifier for some
+% labels. For instance, the embeddings can separate 'future','past' and 'simple
+% sentence' as in Figure X
 % However, it does not work for the rest of groups:
@@
 
-% Whether the degree of transformation can be characterized by SBERT sentence embeddings. 
-% It shows the generalization group achieved most correlation.
-% The more generalized sentences are further from seed sentence. While less generalized sentences (sentence meaning closer to seed sentences) has shorter distance to seed sentences
+% Whether the degree of transformation can be characterized by SBERT sentence
+% embeddings. It shows the generalization group achieved most correlation. The
+% more generalized sentences are further from seed sentence. While less
+% generalized sentences (sentence meaning closer to seed sentences) has shorter
+% distance to seed sentences
 
-\bibliography{anthology}
+\bibliography{bibliography}
 \bibliographystyle{acl_natbib}

### 2023-05-08 paper/paper.tex
Add paper structure and classification paper chapters.

@@
 
-% For proper rendering and hyphenation of words containing Latin characters (including in bib files)
+% For proper rendering and hyphenation of words containing Latin characters
+% (including in bib files)
 \usepackage[T1]{fontenc}
@@
 % \usepackage[T5]{fontenc}
-% See https://www.latex-project.org/help/documentation/encguide.pdf for other character sets
+% See https://www.latex-project.org/help/documentation/encguide.pdf for other
+% character sets
 
@@
 
-% If the title and author information does not fit in the area allocated, uncomment the following
+% If the title and author information does not fit in the area allocated,
+% uncomment the following
 %
@@
 
-\title{Instructions for ACL 2023 Proceedings}
+% Tables
+\usepackage{booktabs}
+\usepackage{multirow}
+
+% Urls
+\usepackage{hyperref}
+
+% For naming embeddings
+\newcommand{\Embed}[1]{\texttt{#1}}
+
+% For naming classifiers
+\newcommand{\Cls}[1]{\texttt{#1}}
+
+% For naming transformations
+\newcommand{\Trans}[1]{\textsl{`#1'}}
+
+\title{Semantic information inside sentence embeddings}
 
@@
     \texttt{leixin.zhang@student.uni-tuebingen.de}
-    }%\\\And
+    }
+\author{David Burian\\
+  Charles University, Prague\\
+  \texttt{david.burian@matfyz.cz}
+  }
   % Second Author \\
@@
 \maketitle
+
 \begin{abstract}
@@
 
-\section{Previous studies}
+\section{Related work}
 
-\section{Methodology}
+% Costra corpus
+Costra corpus~\cite{baranvcikova2020costra}
 
-\section{Evaluating Sentence Embeddings with Gaussian Mixture}
+% SBERT
+SBERT~\cite{reimers2019sentence}
+
+% Doc2vec
+Doc2vec~\cite{le2014distributed}
+
+\section{Experiments}
+
+\subsection{Costra corpus}
+
+% Deep dive into the properties of the datsaet
+%   - transformation terminology
+%   - types and explanations of transformations
+%   - dataset statistics (count of seeds, sentences in total, transformation
+%   distributions)
+
+\subsection{Embeddings}\label{sec:embeddings}
+
+% List the embeddings we are going to work with. Focus only on the core ones,
+% which are used often throughout the tasks.
+We focus mainly on two sets of embeddings: \Embed{doc2vec} predicted by Doc2vec
+model trained on 1M of Czech sentences and \Embed{sbert} predicted by
+pretrained SBERT multilingual model. Additionally we also experimented with
+supervised method using the same SBERT model --- we call this embedding
+\Embed{sbert\_supervised}.
+
+% TODO: doc2vec training
+
+To obtain \Embed{sbert\_supervised} we splitted the corpus into 5 folds. For
+each fold, the model trained on the remaining 4 folds and produced embeddings
+for the one remaining fold. This gave us 5 sets of disjunct embeddings, each
+trained on slightly different data. For training we used softmax loss, which
+classifies sentence and its corresponding seed sentence into the set of all
+possible transformations. This gives the model incentive to produce embeddings
+that are easily distinguishable for different transformations.
+
+\subsection{Tasks}
+
+% High overview of the tasks that were carried out as well as some motivation for
+% doing them. Give the reader a quick overview of the work we've done.
+
+\subsubsection{Probing tasks}
+
+We designed a set of probing tasks which aim to test whether the embeddings
+contain semantic information while ignoring how exactly it is encoded. More
+specifically we ask a classifier to predict the transformation used to obtain
+the given embedding.
+
+The motivation for these tasks is that if the embedding contains semantic
+information (in our case the type of transformation used), it would be
+beneficial for the classifier to find and use this information. This would
+boost its scores and make it standout.
+
+We test two settings: \emph{no-context} and \emph{contextualized}. In
+no-context setting the classifier receives only the sentence embedding itself
+and test which (if any) transformations can be reliably determined without the
+knowledge of the seed sentence. In the contextualized setting we pass both the
+sentence embedding and the embedding of the corresponding seed sentence. This
+enables the classifier to examine the relationships between the transformed and
+seed sentences. The results are presented in Section~\ref{sec:probing_results},
+the complete method is described in detail in
+Appendix~\ref{appendix:probing_method}.
+
+\section{Results}
+
+%Enumeration of the tasks, for each task:
+%  - starBriefly* describe the method used. Citing papers instead of
+%  explaining.
+%  - In case more information is needed (detailed method, implementation
+%  notes), put it in an Appendix.
+%  - Present results. Pinpoint the interesting stuff (in bold?)
+
+\subsection{Probing for transformation prediction}\label{sec:probing_results}
+
+Besides the already described embeddings in Section~\ref{sec:embeddings}, we
+also add their versions with randomly permutated labels within sentences with
+the same seed sentence (named \Embed{mixup\_by\_seed\_}) and several types
+bag-of-words (\emph{BOW}) vectors. The \Embed{mixup\_by\_seed\_} embeddings
+serve as a comparative random baseline. We experimented with two types of BOW
+vectors --- pure bow vector having ones at positions corresponding to words
+included in the given sentence (named \Embed{bow}) and TF-IDF weighted bow
+vectors (named \Embed{tfidf}). Additionally we tried to limit the amount of
+words registered in order to decrease the dimensionality of the vectors,
+however this failed to increase the classifiers' scores and therefore we omit
+them in the rest of this section. Dimensions of all embeddings are presented in
+Table~\ref{tab:embeddings_dim}.
+
+\begin{table}[htp]
+  \centering
+  \begin{tabular}{l  c c}
+    \toprule
+    Embedding & Dimension & Sparseness\\
+    \midrule
+    \Embed{doc2vec} & 512 & 0\%\\
+    \Embed{sbert} & 384 & 0\%\\
+    \Embed{bow} & 8374 & 99.89\%\\
+    \Embed{tfidf} & 8374 & 99.89\%\\
+    \bottomrule
+  \end{tabular}
+  \caption{Dimensionality and sparseness as the percentage of zeros of used
+  embeddings.}\label{tab:embeddings_dim}
+\end{table}
+
+We employ several types of classifiers: Random Forests
+(\Cls{RandomForestClassifier}), Support Vector Machine (\Cls{SVC}) and
+K-Nearest neighbours (\Cls{KNeighboursClassifier}).
+
+In summary, our entire pipeline is as follows:
+\begin{enumerate}
+  \item For each embedding, we consider only a random sample with uniform
+    distribution of transformations and ignore the embeddings of the seed
+    sentences.
+  \item For each embedding and classifier we run a 5-fold cross-validated grid
+    search over several parameters, which gives us the best model for given
+    embedding.
+  \item We train each embedding on the classifier which performed the best
+    using cross-validation.
+\end{enumerate}
+
+More details about the pipeline can be found in
+Appendix~\ref{appendix:probing_method}.
+
+\subsubsection{Contextualized setting}
+
+We tried two methods of contextualization: \emph{concatenating} the two
+embeddings and \emph{subtracting} the seed embedding from the sentence
+embedding. Since the method using difference performed better, we omit the
+concatenation results.
+
+We report the grid search results in
+Figure~\ref{fig:cls_gs_diff_embed_comparison}. The best performing embeddings
+were those produced by SBERT\@. Supervised SBERT embeddings differ
+significantly from the unsupervised ones only in superior performance of
+\Cls{KNeighboursClassifier}. This suggests better local structure in terms of
+the sentence transformation.
+
+\begin{figure}[htp]
+  \centering
+  \includegraphics[width=0.5\textwidth]{figs/cls_gs_diff_embed_comparison.png}
+
+  \caption{Grid search F1 macro averaged scores of contextualized embeddings.
+  The reported scores of \Embed{sbert\_supervised} are the mean scores of all
+  SBERT supervised embeddings.
+  }\label{fig:cls_gs_diff_embed_comparison}
+
+\end{figure}
+
+Moreover the BOW embeddings surpassed \Embed{doc2vec} embedding,
+proving that for prediction of transformation, lexical features are more
+valuable than any information extracted by our trained Doc2vec model.
+
+% TODO?: RandomForest performence on sparse embeddings
+
+In Figure~\ref{fig:cls_gs_diff_labels} we show the performance of best
+classifiers per each label for the most performant dense and lexical
+embeddings: \Embed{sbert} and \Embed{tfidf}. We can identify four
+transformations as being well-predicted having around 0.80 f1-score. However
+the powerful processing of SBERT offered significant advantage to pure lexical
+information only for two of those: \Trans{past} and \Trans{future}. We see even
+bigger difference for \Trans{opposite meaning}, even if it reached just 0.4
+f1-score.
+
+\begin{figure}[htp]
+  \centering
+  \includegraphics[width=0.5\textwidth]{./figs/cls_gs_diff_labels.png}
+
+  \caption{Comparison of F1 score per each transformation of best classifiers
+  for \Embed{sbert} and \Embed{tfidf} embedding in contextualized
+  setting.}\label{fig:cls_gs_diff_labels}
+\end{figure}
+
+In order to examine the most confused transformations we summed confusion
+matrices of the top 3 most performant embeddings. The resulting matrix is shown
+in Figure~\ref{fig:cls_gs_diff_conf_mat}. The most confused pairs of
+transformations are \Trans{minimal change} with \Trans{different meaning} and
+\Trans{nonsense} with \Trans{different meaning}, both in around 40\%.
+Interestingly \Trans{paraphrase} gets recognised almost never, with
+\Trans{simple sentence}, \Trans{formal sentence}, \Trans{nonsense} and
+\Trans{minimal change} having accuracy bellow 20\%.
+
+\begin{figure}[htp]
+  \centering
+  \includegraphics[width=0.5\textwidth]{./figs/cls_gs_diff_conf_mat.png}
+  \caption{Summed confusion matrices of best classifiers for \Embed{sbert},
+  \Embed{sbert\_supervised} and \Embed{tfidf} embeddings. The confusion matrix
+  is row-normalized, whereas the displayed true and predicted distributions are
+  not.}\label{fig:cls_gs_diff_conf_mat}
+\end{figure}
+
+\subsubsection{No-context setting}
+
+Grid search results for no-context setting are displayed in
+Figure~\ref{fig:cls_gs_no_context_embed_comparison}. Comparing these results to
+those in contextualized setting, we see large drop of performance for
+\Embed{sbert} embedding --- almost by 50\%. This shows how much the classifiers
+relied on the information contained in the embedding of seed sentence. This
+drop is not as pronounced for \Embed{sbert\_supervised} embedding, since the
+model producing it is optimized for good scores in precisely this type of
+task.\footnote{For this reason, we avoid comparing \Embed{sbert\_supervised} to
+other embeddings from this point onwards.} On the other hand, the BOW
+embeddings performed almost as well, suggesting they were not really able to
+take advantage of the contextualized setting.
+
+\begin{figure}[htp]
+  \centering
+  \includegraphics[width=0.5\textwidth]{./figs/cls_gs_no_context_embed_comparison.png}
+
+  \caption{Grid search F1 macro averaged scores of sentence embeddings with no
+  context. The reported scores of \Embed{sbert\_supervised} are the mean scores
+  of all SBERT supervised
+  embeddings.}\label{fig:cls_gs_no_context_embed_comparison}
+
+\end{figure}
+
+In Figure~\ref{fig:cls_gs_no_context_labels} we compare performance of best
+classifiers for \Embed{sbert} and \Embed{tfidf} embedding. Whereas the most
+recognised transformations are the same as in the contextualized setting, the
+classifiers performance significantly drops in all cases except for the
+\Trans{ban} and the \Trans{possibility} transformations embedded with TF-IDF.
+This suggests that \Trans{ban} and \Trans{possibility} transformations can be
+easily identified from pure \emph{lexical} features. While SBERT fails to
+encode this information into its embedding, it performs even better when the
+embeddings of a sentence and its corresponding seed sentence are put into one
+context. This phenomena is pronounced mainly for \Trans{different meaning} and
+\Trans{past} transformations, where the inclusion of seed sentence's embedding
+on the input is responsible for over 80\% and 60\% of F1 score of the
+classifier, as can be seen in Figure~\ref{fig:cls_gs_labels_context_diff}.
+Overall SBERT's embeddings benefit from contextualization more than TF-IDF
+vectors.
+
+\begin{figure}[htp]
+  \centering
+  \includegraphics[width=0.5\textwidth]{figs/cls_gs_no_context_labels.png}
+
+  \caption{Comparison of F1 score per each transformation of best classifiers
+  for \Embed{sbert} and \Embed{tfidf} embedding in no-context
+  setting.}\label{fig:cls_gs_no_context_labels}
+
+\end{figure}
+
+\begin{figure}[htp]
+  \centering
+  \includegraphics[width=0.5\textwidth]{figs/cls_gs_labels_context_diff.png}
+
+  \caption{Difference in F1 scores between no-context and contextualized
+  setting as a percentage of contextualized
+  score.}\label{fig:cls_gs_labels_context_diff}.
+
+\end{figure}
+
+\subsection{Evaluating Sentence Embeddings with Gaussian Mixture}
 
@@
 
+\clearpage
+\appendix
+\section{SBERT embeddings}\label{appendix:sbert_embeddings}
+
+For SBERT implementation we used the
+\texttt{sentence\_transformers}\footnote{\url{https://www.sbert.net/}} python
+library.
+
+To produce sentence embeddings with SBERT, we chose pretrained model referred
+to as `\textit{paraphrase-multilingual-MiniLM-L12-v2}'. It offers relatively
+good scores for its size.
+
+For the supervised SBERT sentence embeddings we list the hyperparameters used
+in Table~\ref{tab:sbert_supervised_hparams}. Note that due to the small size of
+training data, the model overfitted quickly. We therefore could afford to train
+only for 4 epochs.
+
+\begin{table}[htp]
+  \centering
+  \begin{tabular}{l c}
+    \toprule
+    Hyperparameter & Value\\
+    \midrule
+    epochs & 4 \\
+    batch size & 8 \\
+    warmup schedule & linear \\
+    warmup steps & 10000 \\
+    optimizer & AdamW \\
+    weight decay & 0.1 \\
+    learning rate & 2e-5 \\
+    \bottomrule
+  \end{tabular}
+
+  \caption{Hyperparameters used to train SBERT for supervised
+  embeddings}\label{tab:sbert_supervised_hparams}
+
+\end{table}
+
+\section{Probing method in detail}\label{appendix:probing_method}
+
+For probing tasks whose results we described in
+Section~\ref{sec:probing_results}, we used
+\texttt{scikit-learn}\footnote{\url{https://scikit-learn.org/stable/}} python
+package.
+
+The grid searched parameters and classifiers are presented in
+Table~\ref{tab:cls_gs_all_params}. We used cross-validated grid search, where
+the performence of a set of parameters is evaluated using cross validation. For
+all cross validations (both within and outside of grid search) we used
+\texttt{StratifiedGroupKFold}, where seed identifiers were used as groups. This
+caused the folds to be stratified as well as sharing as few common seed
+sentences as possible.
+
+\begin{table*}[htp]
+  \centering
+
+  \begin{tabular}{l l l}
+    \toprule
+    Classifier & Hyperparameter & Values searched \\
+    \midrule
+    \multirow{3}{13em}{\Cls{RandomForestClassifier}}
+    & \texttt{n\_estimators} & 50, 100, 200 \\
+    & \texttt{max\_depth} & 2, 5, 25, None \\
+    & \texttt{min\_samples\_split} & 2, 10, 20 \\
+    \midrule
+    \multirow{2}{13em}{\Cls{SVC}*}
+    & \texttt{kernel} & rbf, linear \\
+    & \texttt{gamma} & auto, scale\\
+    \midrule
+    \multirow{2}{13em}{\Cls{KNeighboursClassifier}*}
+    & \texttt{n\_neighbours} & 3, 5, 10\\
+    & \texttt{weights} & uniform, distance\\
+    \bottomrule
+  \end{tabular}
+
+  \caption{Classifiers and hyperparameters grid search for probe tasks. All are
+  named equally as in the \texttt{scikit-learn} python package. *Embeddings
+  were first scaled to have zero mean and unit
+  variance.}\label{tab:cls_gs_all_params}
+
+\end{table*}
+
+We report the best performing set of parameters for each classifier in
+Tables~\ref{tab:cls_gs_best_params_SVC},~\ref{tab:cls_gs_best_params_RandomForestClassifier}~and~\ref{tab:cls_gs_best_params_KNeighborsClassifier}
+
+\begin{table*}[htp]
+  \centering
+  \input{./cls_gs_best_params_SVC_tab.tex}
+
+  \caption{Best parameters for \Cls{SVC}
+  classifier.}\label{tab:cls_gs_best_params_SVC}
+
+\end{table*}
+
+\begin{table*}[htp]
+  \centering
+  \input{./cls_gs_best_params_RandomForestClassifier_tab.tex}
+
+  \caption{Best parameters for \Cls{RandomForestClassifier}
+  classifier.}\label{tab:cls_gs_best_params_RandomForestClassifier}
+
+\end{table*}
+
+\begin{table*}[htp]
+  \centering
+  \input{./cls_gs_best_params_KNeighborsClassifier_tab.tex}
+
+  \caption{Best parameters for \Cls{KNeighboursClassifier}
+  classifier.}\label{tab:cls_gs_best_params_KNeighborsClassifier}
+
+\end{table*}
+
+% chosen parameters
+
 \end{document}

### 2023-05-10 paper/paper.tex
Added a section about Doc2vec training

@@
   }
+ \author{Vojtěch John\\
+ Charles University, Prague\\
+ }
   % Second Author \\
@@
 
-% TODO: doc2vec training
+% TODO: doc2vec training - reread
+For training the Doc2vec model, we have used the implementation of Doc2vec algorithm
+available in the Python package gensim, using a window of size 4 and considering every
+word that occured at all (no pruning of rare words) and the vector size of 512.
+We have also tried training Doc2Vec models with different embedding sizes (2, 3, 10, 100, 256),
+partly so that we could see, ehat impact the size change has on the semantic label predictability, 
+partly (especially if we detected a positive correlation between the embedding size and the semantic label
+predictability) in order to try and get hint as to how prominent the specific semantic features are 
+(i.e. which is the smallest vector size at which given semantic label is predictable from the corresponding embeddings).
 

### 2023-05-11 paper/paper.tex
Update paper.tex

@@
 
+\section{Methodology of Embedding Evaluation}
+
+In addition to visualizing the embeddings of derivation sentences, we employ several approaches to test whether our trained sentence embeddings can accurately characterize the derivation sentence types.
+
+Firstly, we use cosine similarity ......
+
+the Calinski-Harabasz Index, an internal cluster validation test, to pairwise test the derivation clusters and evaluate how well a pair of derivation clusters are separated in space.
+
+Secondly, an unsupervised Gaussian Mixture model is employed to cluster each pair of derivation classes and compute the accuracy of unsupervised cluster labels and with the true labels.
+
+Lastly, we employ several supervised approaches to classify the embeddings and evaluate the extent to which the derivation classes can be captured by the trained classifiers.
+The accuracy scores are compared among classifiers including Multilayer Perceptron, Decision tree and Support Vector Machine.
+
 \subsubsection{Probing tasks}
@@
 
+\section{Distance vectors of derivation class}
+
+This section utilizes the distance vectors between derived sentence embeddings and the seed sentence embedding to predict the derived embedding for another seed.
+We hypothesize that the distance vectors for each derivation class share a similarity and are able to predict the derivation embeddings given the seed embedding.
+For instance, for the derivation class `future', we assume that the following property holds: given two seed sentences  $Seed_i$ and $seed_j$, and one future sentence, $future_i$, we can predict $future_j$.
+
+\begin{equation}
+\label{eq:1}
+ future_j = future_i - seed_i + seed_j
+\end{equation}
+
+To test this hypothesis, we calculate the cosine similarity between the true sentence embeddings and predicted sentence embeddings using Equation \ref{eq:1}.
+We use 80\% of the data to extract and average the distance vectors ($sent_i$-$seed_i$) for each derivation class.
+We then add these distance vectors to the seed vector for the remaining 20\% of the data and compare the cosine similarity between the predicted and true embeddings.
+
+The results are presented in Table \ref{cosine}.
+Most of the derivation classes achieved a cosine similarity score above 0.8 compared with true embeddings, indicating that the predicted vectors closely approximate the true vectors in space, and the distance vectors of derived sentences and seed sentences are relatively stable for each derivation class.
+
+However, the 'generalization' class showed the lowest score, which could be attributed to the fact that each seed contains multiple degrees of transformation for the `generalization' class.
+Some derived sentences are more generalized, while others are less so, and averaging 80\% of them may not be sufficient to capture the features of the remaining 20\% of the data.
+
+\begin{table}[!htp]
+\centering
+\begin{tabular}{ c c }
+\toprule
+\vspace{0.1cm}
+\textbf{class}&\textbf{cosine-sim} \vspace{0.15cm}\\ 
+ \hline \\[-1.2ex]
+\vspace{0.1cm} possibility &    0.936635 \\
+\vspace{0.1cm} past & 0.930636 \\
+\vspace{0.1cm} future &     0.923555 \\
+\vspace{0.1cm} different meaning &    0.913798 \\
+\vspace{0.1cm} nonsense &   0.901452 \\
+\vspace{0.1cm} formal sentence &0.882741 \\
+\vspace{0.1cm} minimal change & 0.866246 \\
+\vspace{0.1cm} ban &  0.845272 \\
+\vspace{0.1cm} paraphrase & 0.81659  \\
+\vspace{0.1cm} nonstandard sentence & 0.812346 \\
+\vspace{0.1cm} simple sentence &0.807523 \\
+\vspace{0.1cm} opposite meaning &     0.748853 \\
+\vspace{0.1cm} generalization & 0.697715 \\
+\toprule
+\end{tabular}
+\caption{Cosine Similarity of predicted and true derivation sentence embeddings}
+\label{cosine}
+\end{table}
 \section{Results}

### 2023-05-11 paper/paper.tex
Add hints how to restructure the paper.

@@
 
+Transformation \Trans{future} ....
+
 \author{Leixin Zhang \\ 
@@
 % Deep dive into the properties of the datsaet
-%   - transformation terminology
+%   - transformation terminology = transformation
+%   - contextualized and distance vector to relation vector
 %   - types and explanations of transformations
@@
 %   distributions)
+
+We define \emph{relation vector} as the ....
 
@@
 % TODO: doc2vec training - reread
-For training the Doc2vec model, we have used the implementation of Doc2vec algorithm
-available in the Python package gensim, using a window of size 4 and considering every
-word that occured at all (no pruning of rare words) and the vector size of 512.
-We have also tried training Doc2Vec models with different embedding sizes (2, 3, 10, 100, 256),
-partly so that we could see, ehat impact the size change has on the semantic label predictability, 
-partly (especially if we detected a positive correlation between the embedding size and the semantic label
-predictability) in order to try and get hint as to how prominent the specific semantic features are 
-(i.e. which is the smallest vector size at which given semantic label is predictable from the corresponding embeddings).
+For training the Doc2vec model, we have used the implementation of Doc2vec
+algorithm available in the Python package gensim, using a window of size 4 and
+considering every word that occured at all (no pruning of rare words) and the
+vector size of 512. We have also tried training Doc2Vec models with different
+embedding sizes (2, 3, 10, 100, 256), partly so that we could see, ehat impact
+the size change has on the semantic label predictability, partly (especially if
+we detected a positive correlation between the embedding size and the semantic
+label predictability) in order to try and get hint as to how prominent the
+specific semantic features are (i.e. which is the smallest vector size at which
+given semantic label is predictable from the corresponding embeddings).
 
@@
 
-\section{Methodology of Embedding Evaluation}
+In addition to visualizing the embeddings of derivation sentences, we employ
+several approaches to test whether our trained sentence embeddings can
+accurately characterize the derivation sentence types.
 
-In addition to visualizing the embeddings of derivation sentences, we employ several approaches to test whether our trained sentence embeddings can accurately characterize the derivation sentence types.
+\subsubsection{Visualization}
+
+\subsubsection{Cosine similarity}
 
@@
 
-the Calinski-Harabasz Index, an internal cluster validation test, to pairwise test the derivation clusters and evaluate how well a pair of derivation clusters are separated in space.
+\subsubsection{CHI index}
 
-Secondly, an unsupervised Gaussian Mixture model is employed to cluster each pair of derivation classes and compute the accuracy of unsupervised cluster labels and with the true labels.
+the Calinski-Harabasz Index, an internal cluster validation test, to pairwise
+test the derivation clusters and evaluate how well a pair of derivation
+clusters are separated in space.
 
-Lastly, we employ several supervised approaches to classify the embeddings and evaluate the extent to which the derivation classes can be captured by the trained classifiers.
-The accuracy scores are compared among classifiers including Multilayer Perceptron, Decision tree and Support Vector Machine.
+\subsubsection{GM}
+
+Secondly, an unsupervised Gaussian Mixture model is employed to cluster each
+pair of derivation classes and compute the accuracy of unsupervised cluster
+labels and with the true labels.
+
+% Lastly, we employ several supervised approaches to classify the embeddings and
+% evaluate the extent to which the derivation classes can be captured by the
+% trained classifiers. The accuracy scores are compared among classifiers
+% including Multilayer Perceptron, Decision tree and Support Vector Machine.
 
@@
 
+\section{Conclusion}
+
+% Summarize our results. Put them to context?
+
 \bibliography{bibliography}

### 2023-05-13 paper/paper.tex
Rewrite probing tasks with new terminology.

@@
 \title{Semantic information inside sentence embeddings}
-
-Transformation \Trans{future} ....
 
@@
 
-% TODO: doc2vec training - reread
+% David: I would maybe skip the part about training shorter embeddings. But
+% that can wait for the paper submit.
 For training the Doc2vec model, we have used the implementation of Doc2vec
@@
 information (in our case the type of transformation used), it would be
-beneficial for the classifier to find and use this information. This would
-boost its scores and make it standout.
+beneficial for the classifier to find it and use it. This would boost its
+scores and make it standout.
 
-We test two settings: \emph{no-context} and \emph{contextualized}. In
-no-context setting the classifier receives only the sentence embedding itself
-and test which (if any) transformations can be reliably determined without the
-knowledge of the seed sentence. In the contextualized setting we pass both the
-sentence embedding and the embedding of the corresponding seed sentence. This
-enables the classifier to examine the relationships between the transformed and
-seed sentences. The results are presented in Section~\ref{sec:probing_results},
-the complete method is described in detail in
-Appendix~\ref{appendix:probing_method}.
+Since transformations define the relationship of two embeddings, we hypothesize
+that classifiers need both the sentence embedding and the seed embedding to
+produce reliable predictions. To test our hypothesis we run the probing tasks
+twice: first classifying transformation vectors\footnote{We also tried
+concatenating the sentence and the seed embeddings. This approach produced
+overall worse results, and therefore we omit them.}, second classifying only
+the sentence embeddings themselves. The results are presented in
+Section~\ref{sec:probing_results}, the complete method is described in detail
+in Appendix~\ref{appendix:probing_method}.
 
@@
 
-\subsubsection{Contextualized setting}
-
-We tried two methods of contextualization: \emph{concatenating} the two
-embeddings and \emph{subtracting} the seed embedding from the sentence
-embedding. Since the method using difference performed better, we omit the
-concatenation results.
+\subsubsection{Classifying transformation vectors}
 
@@
 
-  \caption{Grid search F1 macro averaged scores of contextualized embeddings.
+  \caption{Grid search F1 macro averaged scores for transformation vectors.
   The reported scores of \Embed{sbert\_supervised} are the mean scores of all
@@
 
-% TODO?: RandomForest performence on sparse embeddings
+% TODO?: Should I mention the suprising performance of RandomForest on sparse
+% embeddings?
 
@@
 
-  \caption{Comparison of F1 score per each transformation of best classifiers
-  for \Embed{sbert} and \Embed{tfidf} embedding in contextualized
-  setting.}\label{fig:cls_gs_diff_labels}
+  \caption{F1 score of transformation vectors for each transformation of the
+  best classifiers for given embedding.}\label{fig:cls_gs_diff_labels}
+
 \end{figure}
@@
 
-\subsubsection{No-context setting}
+\subsubsection{Classifying sentence embeddings}
 
-Grid search results for no-context setting are displayed in
+Again, we report plotted grid search results in
 Figure~\ref{fig:cls_gs_no_context_embed_comparison}. Comparing these results to
@@
 relied on the information contained in the embedding of seed sentence. This
-drop is not as pronounced for \Embed{sbert\_supervised} embedding, since the
-model producing it is optimized for good scores in precisely this type of
+drop is not as pronounced for \Embed{sbert\_supervised} embedding, because the
+model producing them is optimized with precisely this type of
 task.\footnote{For this reason, we avoid comparing \Embed{sbert\_supervised} to
@@
 
-  \caption{Grid search F1 macro averaged scores of sentence embeddings with no
-  context. The reported scores of \Embed{sbert\_supervised} are the mean scores
+  \caption{Grid search F1 macro averaged scores of sentence embeddings. The
+  reported scores of \Embed{sbert\_supervised} are the mean scores
   of all SBERT supervised
@@
 classifiers for \Embed{sbert} and \Embed{tfidf} embedding. Whereas the most
-recognised transformations are the same as in the contextualized setting, the
-classifiers performance significantly drops in all cases except for the
-\Trans{ban} and the \Trans{possibility} transformations embedded with TF-IDF.
-This suggests that \Trans{ban} and \Trans{possibility} transformations can be
-easily identified from pure \emph{lexical} features. While SBERT fails to
-encode this information into its embedding, it performs even better when the
-embeddings of a sentence and its corresponding seed sentence are put into one
-context. This phenomena is pronounced mainly for \Trans{different meaning} and
-\Trans{past} transformations, where the inclusion of seed sentence's embedding
-on the input is responsible for over 80\% and 60\% of F1 score of the
-classifier, as can be seen in Figure~\ref{fig:cls_gs_labels_context_diff}.
-Overall SBERT's embeddings benefit from contextualization more than TF-IDF
-vectors.
+recognised transformations are the same as when classifying transformation
+vectors, the classifiers performance significantly drops in all cases except
+for the \Trans{ban} and the \Trans{possibility} transformations embedded with
+TF-IDF. This suggests that \Trans{ban} and \Trans{possibility} transformations
+can be easily identified from pure \emph{lexical} features and based on the
+sentence embedding only. While SBERT fails to encode this information into its
+embedding, it performs even better than TF-IDF when the embeddings of a
+sentence and its corresponding seed sentence are put into one context. As can
+be seen in Figure~\ref{fig:cls_gs_labels_context_diff}, this phenomena is
+pronounced mainly for \Trans{different meaning} and \Trans{past}
+transformations, where the inclusion of seed sentence's embedding on the input
+is responsible for over 80\% and 60\% of F1 score of the classifier. Overall
+SBERT's embeddings benefit from including the seed embedding to the input more
+than TF-IDF vectors.
 
@@
 
-  \caption{Comparison of F1 score per each transformation of best classifiers
-  for \Embed{sbert} and \Embed{tfidf} embedding in no-context
-  setting.}\label{fig:cls_gs_no_context_labels}
+  \caption{F1 scores of sentence embeddings for each transformation of the best
+  classifiers for given embedding.}\label{fig:cls_gs_no_context_labels}
 
@@
 
-  \caption{Difference in F1 scores between no-context and contextualized
-  setting as a percentage of contextualized
-  score.}\label{fig:cls_gs_labels_context_diff}.
+  \caption{Drop in performance of best classifiers for given embedding going
+  from transformation vector to sentence embedding classification. The drop is
+  given as a percentage of score for transformation
+  vectors.}\label{fig:cls_gs_labels_context_diff}.
 

### 2023-05-13 paper/paper.tex
Rewrite probing tasks, part 2

@@
 We define \emph{relation vector} as the ....
+To per
+Costra
 
@@
 Figure~\ref{fig:cls_gs_no_context_embed_comparison}. Comparing these results to
-those in contextualized setting, we see large drop of performance for
+results for probing transformation vectors, we see large drop of performance for
 \Embed{sbert} embedding --- almost by 50\%. This shows how much the classifiers
@@
 embeddings performed almost as well, suggesting they were not really able to
-take advantage of the contextualized setting.
+take advantage of the information inside seed embeddings.
 

### 2023-07-16 clustering/clusters.py
Clustering cleared up

@@ -1,4 +1,10 @@
 #!python3
+
+"""
+How to use:
+python3 clusters.py file_with_embeddings file_for_results
+"""
+
 import sys
 import copy
 from collections import defaultdict
@@ -7,13 +13,17 @@ import sklearn.metrics as sm
 from sklearn.decomposition import PCA
 
 def load_embeddings(filename):
+    """ Load the embeddings from <filename> """
     with open(filename, "r") as r:
         embeddings = [s.strip("\n").split("\t") for s in r]
-        #print(set([x[2] for x in embeddings]))
         embeddings = [(e[:4], [float(n) for n in e[4:]]) for e in embeddings]
         return embeddings
 
 def subtract_seeds(embeddings):
+    """
+    Subtract the seed embeddings from the embeddings 
+    The <embeddings> is a list of tuples ([ID, seed, transformation, sentence], [the embeddings])
+    """
     ebs = defaultdict(list)
     seeds = {}
     for e in embeddings:
@@ -24,9 +34,12 @@ def subtract_seeds(embeddings):
     for k in ebs.keys():
         for e in ebs[k]:
             new_embeddings.append((e[0], [x - y for x, y in zip(e[1], seeds[e[0][1]][1])]))
-    return new_embeddings    
+    return new_embeddings
 
 def transformations(embeddings, trans):
+    """
+    Get the embeddings corresponding to given transformation
+    """
     features = []
     labels = []
     for embedding in embeddings:
@@ -36,87 +49,34 @@ def transformations(embeddings, trans):
     return np.array(features), labels
 
 def pca_fit(inputs, n):
+    """
+    not used - the PCA
+    """
     pca = PCA(n_components = n)
     pca.fit(np.transpose(inputs))
     pca_output = np.transpose(pca.components_)
     return pca_output
 
-def evaluate(trans, pca_n=0,filename=None, embeddings=None):
+def evaluate(trans, pca_n=0, filename=None, embeddings=None):
+    """
+    Rub the Calinski-Harabasz index evaluation
+    """
     if embeddings==None:
         embeddings = load_embeddings(filename)
     f, t = transformations(embeddings, trans)
     if pca_n > 0:
         f = pca_fit(f, pca_n)
-    #return sm.silhouette_score(f,t)
[…]

### 2023-07-16 clustering/clusters.sh

@@ -1,0 +1,2 @@
+#!/bin/bash
+ls $2 | xargs -I % python3 clustering.py $1/% $2

### 2023-07-18 src/accuracy_cancel.py
add new embeddings from LaBSE, phrase-bert

@@ -1,0 +1,11 @@
+import matplotlib.pyplot as plt
+
+from sentence_topology.classification.analysis import ClassifierAnalysisResults
+from sentence_topology.visualization.predictions import draw_confusion_matrix
+doc2vec_results = ClassifierAnalysisResults.load(
+    "./results/best_classifier_analysis-doc2vec-cs-vecsize-512-train-1M-sents.tsv.pkl"
+)
+
+# Un-normalized confusion matrix
+print(doc2vec_results.confusion_matrix)
+print('done')

### 2023-07-29 paper/paper_2023_May/acl2023.tex

\pdfoutput=1
% In particular, the hyperref package requires pdfLaTeX in order to break URLs across lines.

\documentclass[11pt]{article}

% Remove the "review" option to generate the final version.
\usepackage[review]{ACL2023}

% Standard package includes
\usepackage{times}
\usepackage{latexsym}

% For proper rendering and hyphenation of words containing Latin characters
% (including in bib files)
\usepackage[T1]{fontenc}
% For Vietnamese characters
% \usepackage[T5]{fontenc}
% See https://www.latex-project.org/help/documentation/encguide.pdf for other
% character sets

% This assumes your files are encoded as UTF8
\usepackage[utf8]{inputenc}

% This is not strictly necessary, and may be commented out.
% However, it will improve the layout of the manuscript,
% and will typically save some space.
\usepackage{microtype}

% This is also not strictly necessary, and may be commented out.
% However, it will improve the aesthetics of text in
% the typewriter font.
\usepackage{inconsolata}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}

% If the title and author information does not fit in the area allocated,
% uncomment the following
%
%\setlength\titlebox{<dim>}
%
% and set <dim> to something 5cm or larger.

% Tables
\usepackage{booktabs}
\usepackage{multirow}

% Urls
\usepackage{hyperref}

% % For naming embeddings
% \newcommand{ }[1]{\texttt{#1}}

% % For naming classifiers
% \newcommand{\Cls}[1]{\texttt{#1}}

% % For naming transformations
% \newcommand{  }[1]{\textsl{`#1'}}

\title{Unveiling Semantic Information in Sentence Embeddings}

\author{David Burian^1 \And Vojtěch John^1 \And Leixin Zhang^2 \And Ondřej             Bojar^1\\}
 \\
   % Faculty Of Mathematics and Physics, Charles University^1\\
   % Seminar für Sprachwissenschaft, Universität Tübingen^2\\
   %      }

%     Computational Linguistics\\
%     Seminar für Sprachwissenschaft,
%     Universität Tübingen, Germany \\ 
%     \texttt{leixin.zhang@student.uni-tuebingen.de}
%     }
% \author{David Burian\\
%   Charles University, Prague\\
%   \texttt{david.burian@matfyz.cz}
%   }
% \author{Vojtěch John\\
%  Charles University, Prague\\
%  }

\begin{document}
\maketitle

\begin{abstract}

This study evaluates the extent to which semantic information is preserved within sentence embeddings trained with doc2vec and SBERT. 
Specifically, we examined whether sentence embeddings within one class, (e.g., `formal sentence') exhibit similarities and cluster together in high dimensional space, and whether they are separable from embeddings of other classes such as `nonsense'.  Our experiments suggest that embeddings trained by SBERT achieve positive performance, but doc2vec does not show a comparable outcome. Additionally, we observe variations in performance across different classes. Notably, tense-related classes (e.g., `future' and `past') and classes containing specific keywords (e.g., negative imperative) demonstrate higher separability in clustering tests and greater predictability in classification tests.
\end{abstract}

\section{Introduction}

Word embeddings are commonly used as input in deep neural networks. However, representing an entire sentence numerically, specifically, as a vector of a fixed length poses significant challenges. Similarly, there is a need for further research to assess the capacity of sentence embeddings in capturing the semantic information of a sentence.

The challenges arise from both training and assessment of sentence embeddings. Firstly, generating embeddings is not as easy as extracting word embeddings from a text based on the surrounding context. Sentence embeddings can be sparse and less representational. 
On the other hand, evaluating the quality of sentence embeddings is not an easy task. Assessing whether the embeddings capture the meaning of sentences on a large scale requires a human-annotated corpus with well-defined semantic classes or sentence similarity indexes.

In this study,  we convert Czech sentences in Costra corpus into embeddings by using SBERT and  doc2vec models. Costra corpus \cite{baranvcikova2020costra} is a collection of Czech sentences with semantic labels. Each set comprises a `seed' sentence  and other transformation sentences, such as `paraphrase' and `different meaning'. The goal of this study is to assess whether sentence embeddings trained by  SBERT and  doc2vec preserve semantic information in some way and whether embeddings from the same transformation class cluster together and separate from other classes in space.

The paper is structured as follows: Section \ref{sec:related work} provides an overview of the SBERT and doc2vec, along with a detailed introduction to the Costra corpus. Section \ref{sec: embeddings} presents initial experiments of extracting transformation vectors and visualizing them in 2-D graphs. In Section \ref{sec:cosine}, the extracted transformation embeddings are used to predict original SBERT sentence embeddings, and they are compared using cosine similarity. Section \ref{sec:cluster} implements unsupervised methods to pairwise assess the separability of embeddings from other transformation classes. In Section \ref{sec:classification}, supervised methods are employed to train and predict transformation classes, to investigate whether the semantic information (transformation labels) is preserved in sentence embeddings and can be distinguished through classification methods.

\section{Related Work} \label{sec:related work}

This section introduces the two models, SBERT and doc2vec, that were used in our study to generate sentence embeddings. Subsequently, we provide an overview of the Costra corpus \cite{baranvcikova2020costra}, which serves as the dataset for evaluating the performance of the sentence embeddings.

% SBERT
Sentence BERT or SBERT \cite{reimers2019sentence} aggregates pre-trained word embeddings of BERT \cite{devlin-etal-2019-bert} by taking their arithmetic mean and fine-tuning with siamese and triplet network structures \cite{reimers2019sentence}. The model has shown success in many previous studies.\cite{opitz-frank-2022-sbert,9140343}.

% Doc2vec
Doc2vec, also referred to as paragraph vector \cite{le2014distributed} is designed to convert long texts into embedding representations of a fixed length. The randomly initialized paragraph embeddings are updated by concatenating with three subsequent word embeddings and predicting the following word in the text. In our study, we used an additional dataset of 1 million Czech sentences to train the model and use the trained doc2vec to predict sentence embeddings in the Costra corpus.

% Costra corpus

% Deep dive into the properties of the datsaet
%   - transformation terminology = transformation
%   - contextualized and distance vector to relation vector
%   - types and explanations of transformations
%   - dataset statistics (count of seeds, sentences in total, transformation
%   distributions)

Costra corpus \cite{baranvcikova2020costra}, used as an evaluation dataset in our study, consists of 6968 Czech sentences. Among these sentences, there are 126 seed sentences and 13 types of transformation sentences derived from seeds, such as `different meaning', `paraphrase', and `future', as displayed in Table \ref{costra}). Some classes are inherently independent of the seed sentences, such as `past', `future', or `ban'\footnote{`Ban' means negative imperative}. Some are more seed dependent like `paraphrase' or `different meaning'. A seed sentence can contain more than one transformation sentence of the same class. For example, a seed sentence may contain 10 `generalization' sentences of varying transformation degrees (some are more generalized, some are less)

\begin{table}[!htp]
\centering
\begin{tabular}{ c c }
\toprule
\vspace{0.1cm}
\textbf{class}&\textbf{sentence count} \vspace{0.15cm}\\ 
 \hline \\[-1.7ex]

\vspace{0.1cm}seed & 126\\
\vspace{0.1cm}ban & 253\\
\vspace{0.1cm}different meaning & 263\\
\vspace{0.1cm}possibility & 271\\
\vspace{0.1cm}simple sentence & 275\\
\vspace{0.1cm}minimal change & 283\\
\vspace{0.1cm}nonsense & 285\\
\vspace{0.1cm}past & 559\\
\vspace{0.1cm}paraphrase & 585\\
\vspace{0.1cm}future & 637\\
\vspace{0.1cm}opposite meaning & 759\\
\vspace{0.1cm}formal sentence & 783\\
\vspace{0.1cm}generalization & 808\\
\vspace{0.1cm}nonstandard sentence & 1081\\
\toprule
\end{tabular}
\caption{Transformation classes and sentence counts}
\label{costra}
\end{table}

\section{Sentence Embeddings} \label{sec: embeddings}

% List the embeddings we are going to work with. Focus only on the core ones,
% which are used often throughout the tasks.
We mainly examine two types of embeddings: \textbf{sentence embeddings} generated directly from doc2vec and SBERT, and \textbf{transformation embeddings}, which are defined as sentence embeddings subtracted by their corresponding seed embeddings.  For example, we refer to the transformation embedding of `generalization' as:  \textit{generalization $embedding_i$}  -   \textit{$seed_i$}.

In the initial visualization experiments, we observed that sentence embeddings obtained from both doc2vec and SBERT did not exhibit the separability of any transformation classes in the space. Embeddings of all labels mixed together after dimension reduction by PCA and t-SNE projection as in Figure \ref{fig:sbert_original}.

But SBERT transformation embeddings (those subtract their seeds) indicate a positive result: transformation embeddings of the same classes (mainly from tense-related classes)  tend to cluster together as in Figure \ref{fig:sbert_vis}, although this separability is not observed for all 13 classes.

However, the transformation embeddings generated by doc2vec did not demonstrate comparable results to SBERT (see Figure \ref{fig:doc2vec_vis}). Our subsequent experiments further confirmed that the embeddings from doc2vec often failed to surpass the baseline. This observation is consistent with the findings of Lau and Baldwin (\citeyear{lau-baldwin-2016-empirical}), who argue that shorter documents (such as sentences in our case) do not perform as well as longer documents, and the benefits of using doc2vec diminish for shorter paragraphs. As a result, the following experiments will primarily present the results from SBERT transformation embeddings.

\begin{figure}[htp]
    \centering
    \includegraphics[scale=0.3]{figs/sbert_original.png}
    \caption{Visualization of five classes' SBERT sentence embeddings}\label{fig:sbert_original}
\end{figure}
\begin{figure}[htp]
    \centering
    \includegraphics[scale=0.3]{figs/sbert_vis.png}
    \caption{Visualization of five classes' SBERT transformation embeddings (sentence embeddings minus their seeds).}\label{fig:sbert_vis}
\end{figure}
\begin{figure}[htp]
    \centering
    \includegraphics[scale=0.3]{figs/doc2vec.png}
    \caption{Visualization of five classes' doc2vec transformation embeddings (sentence embeddings subtract their seeds) }\label{fig:doc2vec_vis}
\end{figure}

\section{Test 1: Predication Capability of Transformation Embeddings}\label{sec:cosine}

This section tests whether the transformation vectors (sentence embeddings subtract the seed embedding) are capable to predict sentence embeddings for other seeds. We hypothesize that the transformation vectors of the same class share a similarity and are able to predict other sentence embeddings given the seed embedding. For instance, we assume that the following property holds: given two seed sentence embeddings $seed_i$ and $seed_j$, along with the future sentence embedding $future_i$, we can predict future sentence embeddings for $seed_j$ as equation \ref{eq:1} or \ref{eq:2}

\begin{equation}
\label{eq:1}
\small{ future_j = future_i - seed_i + seed_j}
\end{equation}
\begin{equation}
\label{eq:2}
 \small{future_j = Future\ Trans\ Embedding + seed_j}
\end{equation}

To test this hypothesis, We split each class to two parts: 80\% of the data for extracting transformation vectors ($sent_i$-$seed_i$) and calculating their average to predict sentence embeddings for the remaining 20\% data. We assess the performance by measuring the cosine similarity between the predicted sentence embeddings and the true sentence embeddings.

The results presented in Table \ref{cosine} indicate that the majority of the transformation classes achieved a cosine similarity score above 0.8. This score is higher than the benchmark range of 0.72-0.78, which was computed on the dataset with randomized seeds.

These findings suggest that the predicted vectors closely approximate the true vectors in space, and the transformation vectors are relatively consistent for each transformation class.

The result also indicates that, however, the class `generalization' shows a lower cosine similarity score than the benchmark, which could be attributed to the fact that each seed contains multiple degrees of transformation for the `generalization' class. Some sentences are more generalized, while others are less so, and averaging 80\% of them may not be sufficient to predict the embeddings for sentences with varying transformation degrees.

\begin{table}[!htp]
\centering
\begin{tabular}{ c c }
\toprule
\vspace{0.1cm}
\textbf{class}&\textbf{cosine-sim} \vspace{0.15cm}\\ 
 \hline \\[-1.7ex]
\vspace{0.1cm} possibility &    0.936635 \\
\vspace{0.1cm} past & 0.930636 \\
\vspace{0.1cm} future &     0.923555 \\
\vspace{0.1cm} different meaning &    0.913798 \\
\vspace{0.1cm} nonsense &   0.901452 \\
\vspace{0.1cm} formal sentence &0.882741 \\
\vspace{0.1cm} minimal change & 0.866246 \\
\vspace{0.1cm} ban &  0.845272 \\
\vspace{0.1cm} paraphrase & 0.81659  \\
\vspace{0.1cm} nonstandard sentence & 0.812346 \\
\vspace{0.1cm} simple sentence &0.807523 \\
\vspace{0.1cm} opposite meaning &     0.748853 \\
\vspace{0.1cm} generalization & 0.697715 \\ 
\vspace{0.1cm} banchmark (randomized seeds) & 0.72-0.78 \\

\toprule
\end{tabular}
\caption{Cosine Similarity of predicted embeddings and true derivation sentence embeddings}
\label{cosine}
\end{table}

\section{Test 2: Clustering Test of Transformation Embeddings}\label{sec:cluster}
This section analyzes whether the transformation embeddings of the same class cluster together and are separated from other classes in space. Two unsupervised methods are utilized for testing purposes. In Section 5.1, we present an internal cluster validation test using the Calinski-Harabasz Index. In Section 5.2, we use Gaussian Mixture modeling to assess whether transformation classes can be separated with the Gaussian distribution-based method.

\subsection{Calinski-Harabasz Index}

The Calinski-Harabasz Index (Equation \ref{eq:3}) measures the ratio of between-cluster dispersion to inter-cluster dispersion. A higher value signifies well-separated clusters \cite{calinski1974dendrite}. In this study, we conducted pairwise testing of transformation embeddings generated by SBERT to assess the degree of separation between pairs of transformation classes in the vector space.
\begin{equation}
\label{eq:3}
  \scriptsize CH =\left[  {\frac {\sum_{k=1}^{K} n_k  \lVert c_k - c \rVert ^2}{K-1} }\right] /
 \left[ \frac{\sum_{k=1}^{K} \sum_{i=1}^{n_k}  \lVert d_i -c_k  \rVert ^2 }{N-K} \right] 
\end{equation}

We establish a benchmark for the CH Index by mixing up the seeds in the dataset, resulting in a range of 0.37 to 1.05. In our evaluation of SBERT transformation embeddings, we observe that 5 pairs fall below this benchmark (see Table \ref{CH_worst}). This indicates that the embeddings of `paraphrase', `minimal change', and `different meaning' are not separated from their seeds. Similarly, it suggests that the transformation embeddings of these three classes may not be distinguishable from one another.

\begin{table}[!htp]
\centering
\begin{tabular}{c c c}
\toprule
\vspace{0.1cm}
\textbf{class 1} &\textbf{class 2} &\textbf{CH index} \vspace{0.15cm}\\ 
 \hline \\[-1.7ex]
\vspace{0.1cm}paraphrase	& 	seed		& 0.77\\
\vspace{0.1cm}different meaning		&minimal change	&	0.80\\
\vspace{0.1cm}minimal change		&seed	& 0.88\\
\vspace{0.1cm}different meaning	&	seed	&	0.93\\
\vspace{0.1cm}different meaning	&	paraphrase	&	1.04\\
\toprule
\end{tabular}
\caption{5 pairs of classes with CH Index below the baseline}
\label{CH_worst}
\end{table}

\begin{table}[!htp]
\centering
\begin{tabular}{c c c}
\toprule
\vspace{0.1cm}
\textbf{class 1} &\textbf{class 2} & \textbf{CH index} \vspace{0.15cm}\\ 
 \hline \\[-1.7ex]
\vspace{0.1cm} ban	&	past	&	168.04 \\
\vspace{0.1cm} future	& past	&   147.48\\
\vspace{0.1cm} ban	& future	&	145.31 \\
\vspace{0.1cm} ban	& formal sentence	&112.28\\
\vspace{0.1cm} ban	& possibility	&102.32 \\

\toprule
\end{tabular}
\caption{5 pairs of transformation classes with the highest CH Index score}
\label{CH_best}
\end{table}

However, 66\% transformation pairs exhibit a significantly higher CH Index (above 10.00) than the benchmark. The classes with the highest separability from other classes include `ban', `future', `past', and `possibility'. The 5 best pairs are presented in Table \ref{CH_best}.

\subsection{Evaluating Sentence Embeddings with Gaussian Mixture}

This section analyzes the separability of transformation classes using the Gaussian Mixture model. It is a probabilistic model that assumes that each cluster follows a Gaussian (or normal) distribution and estimates the weight of
the density function for each cluster \cite{reynolds2009gaussian,5298967}.

We assume that the transformation embeddings of two distinct classes should achieve high
accuracy with the Gaussian Mixture model if clusters are displayed in a Gaussian distribution
in space and are separable from each other.
We performed pairwise tests to assess the separability of labels and calculated the accuracy (F-score) of the GM clustered labels compared to the true labels.

Our evaluation results show that out of 78 pairs of derivation classes, 7 pairs
achieved an accuracy score above 90\%, 24 pairs above 80\%, and 47 pairs above
60\%, as depicted in Figure~\ref{fig:GM}.

In particular, the class `ban' demonstrates good separability with many other
classes, achieving an accuracy score of 0.982 with `past', 0.980 with `formal
sentence', 0.942 with `minimal change' etc. Additionally, our evaluation results reveal that tense-related classes are generally
well-separated, with an accuracy score of 0.90 for `past' and future' classes,
and an accuracy score of 0.924 for `simple sentence' and `future' classes.

\begin{figure}[htp]
    \centering
    \includegraphics[scale=0.15]{figs/GM.png}
    \caption{Accuracy of class label prediction using Gaussian
    Mixture}\label{fig:GM}
\end{figure}

However, we realized the potential limitation of using this Gaussian Mixture accuracy as a measure of separability. The accuracy score from GM may be underestimated if two clusters are not normally distributed and the boundary between them is fuzzy.

%Enumeration of the tasks, for each task:
%  - starBriefly* describe the method used. Citing papers instead of
%  explaining.
%  - In case more information is needed (detailed method, implementation
%  notes), put it in an Appendix.
%  - Present results. Pinpoint the interesting stuff (in bold?)

\section{Test 3: Classification Tests of Embeddings}\label{sec:classification}

% Besides the already described embeddings in Section~\ref{sec:embeddings} XXX,  and several types
% bag-of-words (   {BOW}) vectors.  We experimented with two types of BOW
% vectors --- pure bow vector having ones at positions corresponding to words
% included in the given sentence (named  {bow}) and TF-IDF weighted bow
% vectors (named  {tfidf}).Dimensions of all embeddings are presented in
% Table~\ref{tab:embeddings_dim}.

In previous experiments, we utilized methods such as visualization, sentence embedding prediction, and clustering validation to assess the separability of classes. However, we observed that some classes may not be well-separated from other classes based on these methods.

This section introduces more sophisticated supervised methods to further evaluate our sentence embeddings.
The classifiers include Random Forests, Support Vector Machine (SVM), and K-Nearest Neighbors (KNN). These methods have their own advantages: Random Forests use specific criteria and feature-based splitting to classify data \cite{breiman2001random, cutler2012random}, while SVM captures the decision boundary between two classes employing a hyperplane \cite{scholkopf1999advances, smola2004tutorial}. KNN uses a distance-based approach and classifies data based on the true labels of nearby data points.

We intend to investigate whether these diverse methods could uncover any semantic information that might be encoded within the sentence embeddings.

In addition to the sentence embeddings from doc2vec and SBERT, we also included supervised SBERT sentence embeddings. They are  based on SBERT embeddings and further trained with MLP to predict class labels. Moreover, two lexical methods, namely bag of words (BoW) and term frequency-inverse document frequency (TF-IDF) weighted BoW, are also added to predicted class labels. The comparison with BoW and TF-IDF embeddings aimed to assess whether the classification performance is influenced by specific words that exclusively appear in a particular class.

% In summary, our entire pipeline is as follows:
% \begin{enumerate}
%   \item For each embedding, we consider only a random sample with uniform
%     distribution of transformations and ignore the embeddings of the seed
%     sentences.
%   \item For each embedding and classifier we run a 5-fold cross-validated grid
%     search over several parameters, which gives us the best model for given
%     embedding.
%   \item We train each embedding on the classifier which performed the best
%     using cross-validation.
% \end{enumerate}

% More details about the pipeline can be found in
% Appendix~\ref{appendix:probing_method}.
% \subsection{Classifying sentence embeddings}
\subsection{Classification Performace of Varying Classifiers}

In this section, we compare three classifiers' performance in predicting transformation classes using two types of input: sentence embeddings (Figure \ref{fig:cls_gs_no_context_embed_comparison}) and transformation embeddings  (Figure \ref{fig:cls_gs_diff_embed_comparison}) in predicting transformation class labels.

\begin{figure}[htp]
  \centering
  \includegraphics[width=0.5\textwidth]{./figs/cls_gs_no_context_embed_comparison.png}
  \caption{Averaged F1-macro scores of sentence embeddings from three classifiers. (The
  reported scores of  {sbert\_supervised} are computed as the mean
  of all SBERT supervised
embeddings.)}\label{fig:cls_gs_no_context_embed_comparison}
\end{figure}
\begin{figure}[htp]
  \centering
  \includegraphics[width=0.5\textwidth]{figs/cls_gs_diff_embed_comparison.png}
  \caption{Average F1-macro scores for transformation embeddings from three classifiers.
  (The reported scores of  {sbert\_supervised} are computed as the mean of all
  SBERT supervised embeddings.
  }\label{fig:cls_gs_diff_embed_comparison}
\end{figure}

By comparing the performance of sentence embeddings (Figure \ref{fig:cls_gs_no_context_embed_comparison}) and transformation embeddings (Figure \ref{fig:cls_gs_diff_embed_comparison}), it can be observed that there is a significant difference in the performance of SBERT sentence embeddings and SBERT transformation embeddings. SBERT sentence embeddings perform even worse than TF-IDF.

This suggests that embeddings that retain seed information introduce additional noise and make it more challenging to predict the correct classes. Subtraction of seed embeddings plays a crucial role in improving the prediction from SBERT embeddings. However, this distinction between sentence embeddings and transformation embeddings is less pronounced when using supervised SBERT embeddings. It suggests that preliminarily fine-tuning SBERT on the current task helps mitigate the impact of such noise.

The results of transformation embeddings in Figure \ref{fig:cls_gs_diff_embed_comparison} indicate that the embeddings of the best performance were generated by SBERT. Supervised SBERT embeddings demonstrated a significant advantage over unsupervised SBERT embeddings only when using k-nearest neighbors. This may reveal that KNN, the distance-based method relying on neighboring data points may not fully capture the relationships of the same class in high-dimensional space. The overall performance of KNN can be enhanced by using supervised trained embeddings as input.

Additionally, our findings suggest that TF-IDF weighted embeddings generally outperform the count-based Bag of Words (BoW) method. In the subsequent sections, we will mainly analyze predictions of SBERT and TF-IDF embeddings using the best-performing classifiers.

\subsection{Accuracy of Prediction of Class Labels}

In this section,  we analyze the prediction accuracy of SBERT and TF-IDF from sentence embeddings (Figure~\ref{fig:cls_gs_no_context_labels}) and transformation embeddings (Figure \ref{fig:cls_gs_diff_labels}).

It is observed that performance from SBERT and TF-IDF display a correlation in both figures. A class that achieves a high F1 score with TF-IDF embeddings tends to have a high score with SBERT too. This finding suggests that the high accuracy score may be attributed to lexical factors. Specifically, the presence of certain words in the sentence may make the transformation classes more predictable. These results may also indicate that SBERT heavily relies on specific lexical information when encoding semantic information.

\begin{figure}[htp]
  \centering
  \includegraphics[width=0.5\textwidth]{figs/cls_gs_no_context_labels.png}
  \caption{F1 scores of sentence embeddings from the best classifier}\label{fig:cls_gs_no_context_labels}
\end{figure}
\begin{figure}[htp]
  \centering
  \includegraphics[width=0.5\textwidth]{./figs/cls_gs_diff_labels.png}

  \caption{F1 score of transformation embeddings from the
  best classifiers}\label{fig:cls_gs_diff_labels}
\end{figure}

Moreover, the results indicate transformation embeddings generally perform better than sentence embeddings, suggesting subtracting the corresponding seed embeddings improves the prediction accuracy for many classes. This tendency is pronounced for `different meaning', `future' and `past' (Figure \ref{fig:cls_gs_labels_context_diff}). 
\begin{figure}[htp]
  \centering
  \includegraphics[width=0.5\textwidth]{figs/cls_gs_labels_context_diff.png}

  \caption{Updated performance by using transformation embeddings, compared with sentence embeddings from the best classifiers.}
 % The result is represented as a percentage of the score for transformation
 %  vectors.}
 \label{fig:cls_gs_labels_context_diff}.
\end{figure}
\begin{figure}[htp]
  \centering
  \includegraphics[width=0.5\textwidth]{./figs/cls_gs_diff_conf_mat.png}
  \caption{confusion matrix by summing up prediction from three sets of sentence embeddings (sbert, sbert\_supervised and TF-IDF) obtained from the best classifiers. The confusion matrix
  is normalized for each row.}
  %whereas the displayed true and predicted distributions are
  %not.}
  \label{fig:cls_gs_diff_conf_mat}
\end{figure}

In order to assess the difficulty of predicting specific transformation classes, we present a confusion matrix (Figure~\ref{fig:cls_gs_diff_conf_mat}) (based on the sum of predictions from the three best embeddings). The pairs with the least separability are `minimal change' vs. `different meaning' and `nonsense' vs. `different meaning'. Notably, the class `paraphrase' is frequently mispredicted (with an accuracy below 20\%.), and often gets mixed up with `simple sentence', `formal sentence', `nonsense', and `minimal change'.

% the classifiers' performance significantly drops in all cases except
% for the `ban' and the `possibility' transformations embedded with
% TF-IDF. This suggests that    `ban' and   `possibility' transformations
% can be easily identified from pure    {lexical} features and based on the
% sentence embedding only. the most seed-independent classes, followed by future and past. 
% While SBERT fails to encode this information into its
% embedding, it performs even better than TF-IDF when the embeddings of a
% sentence and its corresponding seed sentence are put into one context. As can
% be seen in Figure~\ref{fig:cls_gs_labels_context_diff}, this phenomena is
% pronounced mainly for   {different meaning} and   {past}
% transformations, where the inclusion of seed sentence's embedding on the input
% is responsible for over 80\% and 60\% of F1 score of the classifier. Overall
% SBERT's embeddings benefit from including the seed embedding to the input more
% than TF-IDF vectors.

% Moreover the BOW embeddings surpassed  doc2vec embedding,
% proving that for prediction of transformation, lexical features are more
% valuable than any information extracted by our trained Doc2vec model.

% TODO?: Should I mention the suprising performance of RandomForest on sparse
% embeddings?
 % In Figure~\ref{fig:cls_gs_diff_labels} we show the best performance of each class from  SBERT and pure lexical embedding  {tfidf}. Four
% transformation classes are being well-predicted by SBERT with the f1-score of around 0.80.

\section{Conclusion}

Our study analyzed two sets of sentence embeddings: SBERT and doc2vec, and test tested them using the Costra corpus. We found that SBERT performed well in most tests, while the results from doc2vec were unsatisfactory.

Section \ref{sec:cosine} suggests that SBERT transformation embeddings (obtained by subtracting the seed embedding from the sentence embedding) from the corpus were capable of inferring the same transformation type for other seed sentences. However, the effectiveness of the predictions may be influenced by the degree of transformation.

Regarding the separability of transformation classes, we conducted internal validity tests, unsupervised prediction, and supervised prediction of transformation labels. The results indicated that transformation embeddings generally perform better than original sentence embeddings. The experiment results also suggest that some classes exhibited higher separability than others, particularly those related to tense and transformation classes of `ban' and `possibility'. However, classes such as `minimal change', `paraphrase', and `different meaning' tend to be mixed up with each other. This tendency may also be consistent with the human judgment of transformation labels. In other words, these classes can be inherently difficult to distinguish.

One class that proved challenging for the embeddings to recognize, but not for humans is `non-sense'. This highlights the limitation of our sentence embeddings: although SBERT captures the meaning of individual words in the sentence, they lack the overall intelligence to understand the meaning of a sentence. The same finding is also supported by the positive correlation between SBERT and TF-IDF embeddings' performance, suggesting that sentence BERT embeddings, obtained by averaging pre-trained SBERT embeddings rely on the information of individual words. It also implies that inferring meaning beyond the word level is not be feasible using SBERT embeddings.

\bibliography{anthology}
\bibliographystyle{acl_natbib}

\clearpage
\appendix
\section{Other embeddings in the experiments}
\begin{table}[htp]
  \centering
  \begin{tabular}{l  c c}
    \toprule
    Embedding & Dimension & Sparseness\\
    \midrule
     doc2vec & 512 & 0\%\\
     SBERT & 384 & 0\%\\
     {bow} & 8374 & 99.89\%\\
     {tfidf} & 8374 & 99.89\%\\
    \bottomrule
  \end{tabular}
  \caption{Dimensionality and sparseness as the percentage of zeros of used
  embeddings.}\label{tab:embeddings_dim}
\end{table}
\section{SBERT embeddings}\label{appendix:sbert_embeddings}

For SBERT implementation we used the
\texttt{sentence\_transformers}\footnote{\url{https://www.sbert.net/}} python
library.

To produce sentence embeddings with SBERT, we chose pretrained model referred
to as `\textit{paraphrase-multilingual-MiniLM-L12-v2}'. It offers relatively
good scores for its size.

For the supervised SBERT sentence embeddings we list the hyperparameters used
in Table~\ref{tab:sbert_supervised_hparams}. Note that due to the small size of
training data, the model overfitted quickly. We therefore could afford to train
only for 4 epochs.

\begin{table}[htp]
  \centering
  \begin{tabular}{l c}
    \toprule
    Hyperparameter & Value\\
    \midrule
    epochs & 4 \\
    batch size & 8 \\
    warmup schedule & linear \\
    warmup steps & 10000 \\
    optimizer & AdamW \\
    weight decay & 0.1 \\
    learning rate & 2e-5 \\
    \bottomrule
  \end{tabular}

  \caption{Hyperparameters used to train SBERT for supervised
  embeddings}\label{tab:sbert_supervised_hparams}

\end{table}

\section{Probing method in detail}\label{appendix:probing_method}

For probing tasks whose results we described in
Section~\ref{sec:probing_results}, we used
\texttt{scikit-learn}\footnote{\url{https://scikit-learn.org/stable/}} python
package.

The grid searched parameters and classifiers are presented in
Table~\ref{tab:cls_gs_all_params}. We used cross-validated grid search, where
the performence of a set of parameters is evaluated using cross validation. For
all cross validations (both within and outside of grid search) we used
\texttt{StratifiedGroupKFold}, where seed identifiers were used as groups. This
caused the folds to be stratified as well as sharing as few common seed
sentences as possible.

\begin{table*}[htp]
  \centering

  \begin{tabular}{l l l}
    \toprule
    Classifier & Hyperparameter & Values searched \\
    \midrule
    \multirow{3}{13em}{\Cls{RandomForestClassifier}}
    & \texttt{n\_estimators} & 50, 100, 200 \\
    & \texttt{max\_depth} & 2, 5, 25, None \\
    & \texttt{min\_samples\_split} & 2, 10, 20 \\
    \midrule
    \multirow{2}{13em}{\Cls{SVC}*}
    & \texttt{kernel} & rbf, linear \\
    & \texttt{gamma} & auto, scale\\
    \midrule
    \multirow{2}{13em}{\Cls{KNeighboursClassifier}*}
    & \texttt{n\_neighbours} & 3, 5, 10\\
    & \texttt{weights} & uniform, distance\\
    \bottomrule
  \end{tabular}

  \caption{Classifiers and hyperparameters grid search for probe tasks. All are
  named equally as in the \texttt{scikit-learn} python package. *Embeddings
  were first scaled to have zero mean and unit
  variance.}\label{tab:cls_gs_all_params}

\end{table*}

We report the best performing set of parameters for each classifier in
Tables~\ref{tab:cls_gs_best_params_SVC},~\ref{tab:cls_gs_best_params_RandomForestClassifier}~and~\ref{tab:cls_gs_best_params_KNeighborsClassifier}

\begin{table*}[htp]
  \centering
  \input{./cls_gs_best_params_SVC_tab.tex}

  \caption{Best parameters for \Cls{SVC}
  classifier.}\label{tab:cls_gs_best_params_SVC}

\end{table*}

\begin{table*}[htp]
  \centering
  \input{./cls_gs_best_params_RandomForestClassifier_tab.tex}

  \caption{Best parameters for \Cls{RandomForestClassifier}
  classifier.}\label{tab:cls_gs_best_params_RandomForestClassifier}

\end{table*}

\begin{table*}[htp]
  \centering
  \input{./cls_gs_best_params_KNeighborsClassifier_tab.tex}

  \caption{Best parameters for \Cls{KNeighboursClassifier}
  classifier.}\label{tab:cls_gs_best_params_KNeighborsClassifier}

\end{table*}

% chosen parameters

\end{document}
